---
title: "Introduction to Quarto — Lab"
subtitle: "RaukR 2026 · Day 1"
author: ""
date: ""
format:
  html:
    toc-location: left   # free the right margin for the "Quarto docs" pointers
---

::: {.callout-note}
## Scope
Build one penguins document with the authoring extras, then **cite it** and render it as a
**branded Typst PDF**. You will use one document for two challenges, with a break between them.
The result follows the same path from an analysis to a submission-ready report. Everything runs on
**base-R `datasets::penguins`** (R ≥ 4.5), so there is no data to download.

**Before you start:** work through the **[Setup page](../../setup.qmd)** if you haven't. The
`00-check-setup.R` script is at the top of the exercises download (not inside `day1-intro/`). It
confirms R, Quarto, and the packages this lab uses.
:::

::: {.callout-tip appearance="simple"}
You are working in the **`day1-intro/`** folder from the exercises download. Open that folder in
your editor, or make it your working directory in R (see the
[Setup page](../../setup.qmd#get-the-materials)). Everything you render is written next to
your files. You start from a file that is already there, `authoring-starter.qmd`, and the other files
are used later, each at the step that names it (the folder's README lists them).
:::

```{r}
#| label: setup
#| include: false
library(dplyr)
library(ggplot2)
library(gt)
library(ggokabeito)   # color-blind-safe (Okabe-Ito) scale

data(penguins)
penguins <- penguins |> filter(!is.na(bill_len), !is.na(bill_dep))
```

::: {.callout-tip collapse="true"}
## Coming from R Markdown? (optional)
You do **not** need to convert anything to work natively in Quarto. Start with `.qmd` directly.
But if you have an `.Rmd` you want to bring over:

- it renders **as-is** (Quarto reads `.Rmd` too).
- cell options move from the chunk header to the `#|` "hash-pipe" (one YAML option per line).
- `knitr::convert_chunk_header("doc.Rmd", output = "doc.qmd")` does the mechanical part.

The rest of this lab uses native `.qmd`.
:::

## Authoring Challenge

> **Goal:** a penguins report with a cross-referenced **figure**, a cross-referenced **table**, and a
> **margin** element, rendered to HTML.

**Starting point.** Open **`day1-intro/authoring-starter.qmd`** and save it as **`my-report.qmd`**,
in that same folder (next to `references.bib` / `apa.csl` / `_brand.yml`, so Part 2's paths and
branding resolve later). Render it once to confirm it works, then re-render after every step below.

The starter is already a working report, but it has no captions, cross-references, or alt text. Most
steps require no R changes. The accessibility step makes three small changes to the plot.

Most steps use cell options. A cross-reference has two ends: a labeled target and an `@label` in the
prose that points at it. For figures and tables, add a cell label with the right prefix (`fig-` or
`tbl-`) and a caption. (A cell is what R Markdown calls a chunk.) An equation label (`eq-`) attaches
to the closing `$$` instead. Layout also uses a cell option. Each step links to the relevant
documentation.

Missing the download? Take the source directly:

<a class="btn btn-primary btn-sm" href="https://raw.githubusercontent.com/cderv/raukr-2026-quarto-exercises/main/day1-intro/authoring-starter.qmd" role="button" target="_blank">{{< fa download >}} authoring-starter.qmd</a>

### Cross-reference the figure

**Goal.** Turn the scatter plot into a **numbered figure with a caption**, and refer to that number
from the prose. Two edits: an option on the **cell**, and a reference in the **body text** above it.

::: {.column-margin}
{{< fa book-open >}} **Quarto docs**

[Cross-references](https://quarto.org/docs/authoring/cross-references.html)
:::

::: {.callout-note appearance="simple" icon=false}
## {{< fa circle-check >}} You should see
A caption **Figure 1: …** under the plot, and the sentence above it opening with a live **Figure 1**
link.
:::

::: {.callout-tip collapse="true"}
## Hint
Which cell options tell Quarto to number this figure, and how do you refer to it in the text?

A cell becomes cross-referenceable when its label starts with a known prefix (`fig-`
for a figure) and is unique in the document. The caption is a second cell option. Label, caption and
the `@fig-something` reference work together: a reference with no matching label renders as
`?@fig-something`.
:::

::: {.callout-note collapse="true"}
## Solution
On the plot cell:

```r
#| label: fig-bill
#| fig-cap: "Bill length versus depth, colored by species."
```

And the sentence above the cell, rewritten to point at it:

``` markdown
@fig-bill shows that within a species the bill has a characteristic shape, so the species form
clusters.
```
:::

### Cross-reference the table

**Goal.** Apply the same method to the summary table under **Mean measurements**: a numbered, captioned
**Table 1**, introduced by a sentence that links to it. The `gt` call is already written. That
section has no prose yet, so add the sentence.

::: {.column-margin}
{{< fa book-open >}} **Quarto docs**

[Cross-references for tables](https://quarto.org/docs/authoring/cross-references.html#tables)
:::

::: {.callout-note appearance="simple" icon=false}
## {{< fa circle-check >}} You should see
**Table 1: …** above the table (table captions go above, figure captions below), and a sentence
introducing it with a live **Table 1** link.
:::

::: {.callout-tip collapse="true"}
## Hint
Figures and tables are numbered in separate sequences. Which label prefix tells Quarto this one is
a table?

Use `#| label: tbl-summary` and `#| tbl-cap:` on the cell, then `@tbl-summary` in the prose.
The prefix decides both the sequence and the word ("Figure", "Table") Quarto prints.
:::

::: {.callout-note collapse="true"}
## Solution
On the `gt` cell:

```r
#| label: tbl-summary
#| tbl-cap: "Mean measurements per species."
```

And above it:

``` markdown
@tbl-summary reports the mean bill measurements and body mass per species.
```
:::

### Move the counts into the margin

**Goal.** Put the species-counts table in the **right margin**, beside the paragraph that mentions
it, instead of in the main text column. The `column:` cell option controls this. The linked
documentation lists its values.

::: {.column-margin}
{{< fa book-open >}} **Quarto docs**

[Article layout](https://quarto.org/docs/authoring/article-layout.html)
:::

::: {.callout-note appearance="simple" icon=false}
## {{< fa circle-check >}} You should see
The counts table in the right-hand margin, level with "The species are not evenly represented:", and
the main text column unchanged in width.
:::

::: {.callout-tip collapse="true"}
## Hint
The R code does not change. Add `#| column: margin` to the cell. That option also takes `body`,
`body-outset`, `page` and `screen`. For prose rather than cell output, use a
`::: {.column-margin}` div.
:::

::: {.callout-note collapse="true"}
## Solution
On the counts cell:

```r
#| column: margin
```
:::

### Make the figure accessible

**Goal.** Make the figure readable by someone who cannot distinguish the colors, and by someone
using a screen reader.

The plot separates species by **color alone**, in ggplot2's default palette, which is not
color-blind-safe. This is the one step that edits the R: three changes to the plot, one option on the
cell. `ggokabeito` is already loaded in the setup cell and supplies `scale_*_okabe_ito()`.

::: {.column-margin}
{{< fa book-open >}} **Quarto docs**

[HTML accessibility](https://quarto.org/docs/output-formats/html-accessibility.html) · [Figure alt text](https://quarto.org/docs/authoring/figures.html#alt-text)
:::

::: {.callout-note appearance="simple" icon=false}
## {{< fa circle-check >}} You should see
Each species drawn with **its own shape as well as its own color**, in the Okabe-Ito palette (orange,
blue, green rather than ggplot2's salmon, green and blue), and one legend for both color and shape.
If you get two stacked legends instead, give the shape legend the same title as the color legend.
The alt text is not visible on the page: to check it, view the page source and confirm the `<img>`
tag carries your `alt="…"`. The finished figure looks like this:
:::

```{r}
#| label: fig-target
#| echo: false
#| fig-width: 6
#| fig-height: 3.4
#| fig-cap: "Target: bill length versus depth by species."
#| fig-alt: >-
#|   Scatter plot of bill depth against bill length for three penguin species in distinct colors.
#|   The species form largely separate clusters.
ggplot(penguins, aes(bill_len, bill_dep, color = species, shape = species)) +
  geom_point(alpha = 0.8) +
  scale_color_okabe_ito() +
  labs(x = "Bill length (mm)", y = "Bill depth (mm)") +
  theme_minimal(base_size = 12)
```

::: {.callout-tip collapse="true"}
## Hint
Someone who cannot separate the three colors needs a second way to tell the species apart. What else
can `aes()` map? Where does the plot description go when the other figure settings are cell options?

Map `shape = species` alongside `color = species` in `aes()`, and add
`scale_color_okabe_ito()`. The description is the cell option `#| fig-alt:`. It is not the caption:
the alt text stands in for the image, so it describes what the plot *shows*, not what it *is*.
:::

::: {.callout-note collapse="true"}
## Solution
On the plot cell, next to the label and caption:

```r
#| fig-alt: >-
#|   Scatter plot of bill depth against bill length for penguins of three species, each drawn in a
#|   distinct color and shape. The three species form largely separate clusters.
```

And in the plot itself:

```r
ggplot(penguins, aes(bill_len, bill_dep, color = species, shape = species)) +
  geom_point(alpha = 0.8) +
  scale_color_okabe_ito() +
  labs(x = "Bill length (mm)", y = "Bill depth (mm)", color = "Species", shape = "Species") +
  theme_minimal(base_size = 12)
```

Naming both `color` and `shape` in `labs()` merges them into a single legend.
:::

::: {.callout-tip icon=false}
## {{< fa universal-access >}} Optional accessibility check
Your figure now has alt text and redundant color/shape encoding. To check the whole page, Quarto can
run axe-core in the browser. See
[HTML Accessibility](https://quarto.org/docs/output-formats/html-accessibility.html).

Your header still has `format: html` on one line. The option belongs under that format, so expand it
into a block:

``` yaml
format:
  html:
    axe:
      output: document
```

Re-render, and a panel appears at the bottom of the page. With nothing to report it reads **No
axe-core violations found.** Otherwise each violation is listed with its impact, and hovering a
selector highlights the element it points to.

Remove the `axe:` option before you publish.
:::

### Add a numbered equation *(optional)*

**Goal.** Add the bill-shape ratio (bill length divided by bill depth) as a **numbered display
equation**, and refer to it from the prose.

One caution: inside `$$…$$`, `bill_len` can render as a subscript, so escape the underscore as
`bill\_len`.

::: {.column-margin}
{{< fa book-open >}} **Quarto docs**

[Equations](https://quarto.org/docs/authoring/cross-references.html#equations)
:::

::: {.callout-note appearance="simple" icon=false}
## {{< fa circle-check >}} You should see
The formula centered on its own line with **(1)** at the right margin, and the sentence above it
linking to that number.
:::

::: {.callout-tip collapse="true"}
## Hint
Display math is not a cell, so the label cannot be a cell option. Where can you attach it?

The label is an attribute on the closing `$$`, written `{#eq-ratio}`. The `eq-` prefix
makes it a numbered equation, and `@eq-ratio` links to it like `@fig-` and `@tbl-`.
:::

::: {.callout-note collapse="true"}
## Solution
``` markdown
A compact descriptor of bill shape is the length-to-depth ratio (@eq-ratio):

$$
\text{ratio} = \frac{\text{bill\_len}}{\text{bill\_dep}}
$$ {#eq-ratio}
```
:::

::: {.callout-note appearance="simple" icon=false}
## {{< fa circle-check >}} Checkpoint: the whole document
A numbered **Figure 1** (the bill scatter, three separate clusters), a numbered **Table 1**, and the
counts in the right margin. `@fig-bill` and `@tbl-summary` render as live links, with no `?@` markers
anywhere. If you completed the optional step, you also have a numbered **Equation 1**.
:::

**`authoring-checkpoint.qmd` is this same report with every step above applied.** It is already in
`day1-intro/`. Compare your work against it, or, if you did not finish, save a copy of it as
`my-report.qmd` and continue into Part 2 from there. You can also take the source:

<a class="btn btn-primary btn-sm" href="https://raw.githubusercontent.com/cderv/raukr-2026-quarto-exercises/main/day1-intro/authoring-checkpoint.qmd" role="button" target="_blank">{{< fa download >}} authoring-checkpoint.qmd</a>

## Citations Challenge

> **Goal:** cite the report, then render the *same document* as a **branded Typst PDF**.

::: {.callout-note}
## Starting point
If you finished the Authoring Challenge, continue in **`my-report.qmd`**. Otherwise:

1. Open **`day1-intro/authoring-checkpoint.qmd`** (the completed Part-1 report).
2. Save a copy as **`my-report.qmd`** in `day1-intro/`, next to `references.bib`, `apa.csl`, and
   `_brand.yml`.
3. Continue in that copy.

The instructions below use `my-report.qmd` for both routes.
:::

:::: {.callout-note icon=false}
## {{< fa clipboard-list >}} Tasks

::: {.column-margin}
{{< fa book-open >}} **Quarto docs**

[Citations](https://quarto.org/docs/authoring/citations.html) · [Typst](https://quarto.org/docs/output-formats/typst.html) · [Brand](https://quarto.org/docs/authoring/brand.html)
:::

1. Point the document header at the shared bibliography and citation style:
   ```yaml
   bibliography: references.bib
   csl: apa.csl
   ```
2. Cite the two sources the opening paragraph already credits, without rewriting it. The sentence
   ending *"…collected at Palmer Station, Antarctica."* takes the data citation before the period
   (`Antarctica [@gorman2014].`). In the next sentence, put `[@horst2020]` immediately after
   **palmerpenguins**. The rest of that sentence concerns base R, so do not put the citation at the
   end. Don't paste a second copy of either sentence.
3. Add a **References** section where the list should appear:
   ```markdown
   ## References {.unnumbered}

   ::: {#refs}
   :::
   ```
4. Give it a real **title block** so the PDF reads as a paper. **Replace** your document's
   `author:` line (or add one) with:
   ```yaml
   author:
     - name: Your Name
       affiliation: Your Lab, Your University
   ```
5. Render to HTML and confirm that both citations resolve as *(Gorman et al., 2014)* and
   *(Horst et al., 2020)*, with a reference list holding both entries.
6. Switch the output to **Typst** and render a branded PDF. The branding comes from `_brand.yml`,
   which Quarto reads next to the document you render, so keep `my-report.qmd` in
   `day1-intro/`. You may see `unknown font family` warnings on the first render. See Troubleshooting
   if the final PDF still uses a serif font. Two routes:
   - **CLI** — leave the header as-is and force the target:
     ```bash
     quarto render my-report.qmd --to typst
     ```
   - **From the editor** — set `format: typst` in the YAML header first, *then* render (RStudio's
     **Render** button, or **Quarto: Render Document** in Positron and VS Code). The editor renders
     whatever `format:` is declared, so while the header still says `format: html` you get HTML,
     **not** a PDF.

   The PDF is written **next to its source** in `day1-intro/`.
::::

::: {.callout-note appearance="simple" icon=false}
## {{< fa circle-check >}} You should see
In-text **(Gorman et al., 2014)** and **(Horst et al., 2020)**, a **References** section formatted in
APA with both entries, and the same citations in the Typst PDF, typeset with the RaukR palette. No
`?@gorman2014`, `?@horst2020` or `[?]` markers anywhere.

The formats place the bibliography differently. In HTML, `::: {#refs}` positions it. In the Typst
PDF, the **References** heading stays empty. Typst puts the entries at the end under its own
**Bibliography** heading, after the Session appendix.

**Check the branding applied.** An unbranded PDF builds without any error, so check the **title and
headings**. A branded PDF uses **Albert Sans**. An unbranded PDF uses Typst's default serif. Table 1
remains serif in both because of the `unknown font family` issue described in Troubleshooting. Use
the headings for this check. Serif headings mean `_brand.yml` isn't next to your `.qmd`.
:::

::: {.callout-tip collapse="true"}
## Hint
- The `@key` in `[@gorman2014]` must match a key in `references.bib` **exactly**.
- `csl:` sets the style (here APA). Swap the file to restyle every citation at once.
- Typst reads the `.bib` itself, so keep bibliography fields plain (avoid LaTeX macros like
  `\emph{}` in titles).
- The first Typst render requires a network connection while Quarto downloads the brand's Google
  fonts.
:::

:::: {.callout-note collapse="true"}
## Solution: the header and citations

In the YAML header:

``` yaml
bibliography: references.bib
csl: apa.csl
```

In the prose, append each citation to the sentence that credits its source:

``` markdown
… collected at Palmer Station, Antarctica [@gorman2014]. The data are available through the
**palmerpenguins** package [@horst2020] and are now also included in base R's `datasets` package.
```

A References section (the list renders inside the `#refs` div):

``` markdown
## References {.unnumbered}

::: {#refs}
:::
```

For the branded PDF (`day1-intro/_brand.yml` styles it), either force the target from the CLI:

``` bash
quarto render my-report.qmd --to typst
```

or set `format: typst` in the header and render from the editor.
::::

**`solutions/day1/penguins-report.qmd` is the complete Day-1 report:** the starter with every step of
both challenges applied. Try the challenge first, then open it to compare, or take the source:

<a class="btn btn-primary btn-sm" href="https://raw.githubusercontent.com/cderv/raukr-2026-quarto-exercises/main/solutions/day1/penguins-report.qmd" role="button" target="_blank">{{< fa download >}} penguins-report.qmd</a>

For a more advanced branded PDF example, render `day1-intro/sample-typst.qmd`. It also styles its
table and plot in R and sets Typst page options. Those additions are outside this exercise.

## Bonus — one report per species (optional) {#sec-parameters}

> **Goal:** turn a report into a *parameterized* one: a single source that renders a **separate
> report per species** by passing the species at render time. This section is optional. Skip it if
> time is short.

A **parameter** is a value declared in the YAML that your code can read as `params$…`. Change the
value at render time and the *same document* produces a different report. You do not need separate
source files.

::: {.callout-note}
## Starting point
Open **`day1-intro/parameters-starter.qmd`**. It contains the report structure, data, and plot. Your
task is to make the species an input to the document. A complete reference is available at
**`solutions/day1/penguins-by-species.qmd`**.
:::

:::: {.callout-note icon=false}
## {{< fa clipboard-list >}} Tasks

::: {.column-margin}
{{< fa book-open >}} **Quarto docs**

[Parameters](https://quarto.org/docs/computations/parameters.html)
:::

1. Declare a parameter with a default in the header:
   ```yaml
   params:
     species: Gentoo
   ```
2. **Use it to filter** in the setup cell. `params$species` is a value. Keep the existing missing-value
   checks and add the species condition:
   ```r
   one <- penguins |> filter(species == params$species, !is.na(bill_len), !is.na(bill_dep))
   ```
3. **Use the parameter in the report.** Three places, marked `TODO 3a`, `3b` and `3c` in the file.
   Replace the placeholder species in the opening sentence with inline code
   (`` `{{r}} params$species` ``), and name the species in the closing sentence the same way. Make the
   caption and alt text track the parameter with `!expr`. Cell options are YAML, so the `!expr` tag
   tells Quarto to evaluate the R expression:
   ```yaml
   #| fig-cap: !expr paste0("Bill length versus depth for ", params$species, " penguins.")
   #| fig-alt: !expr paste0("Scatter plot of bill depth against bill length for ", params$species, " penguins.")
   ```
4. **Render a different species from the CLI** with `-P name:value`. Run this from inside
   `day1-intro/`:
   ```bash
   quarto render parameters-starter.qmd -P species:Adelie
   ```
   With no `-P`, the default (`Gentoo`) is used.
::::

::: {.callout-tip}
## Optional: dynamic headings and standalone files
Replace `Species` in the heading with inline code so it shows the selected species:

```markdown
## `{{r}} params$species` at a glance
```

To keep two named reports, render each as a self-contained HTML file:

```bash
quarto render parameters-starter.qmd -P species:Adelie    --output adelie.html    --embed-resources
quarto render parameters-starter.qmd -P species:Chinstrap --output chinstrap.html --embed-resources
```

`--output` prevents the second render from overwriting the first. `--embed-resources` puts each
plot inside its HTML file, so the reports do not share a figure-support folder.
:::

::: {.callout-note appearance="simple" icon=false}
## {{< fa circle-check >}} You should see
Rendering with `-P species:Adelie` gives a report whose sentences, figure caption, alt text, point
count, and plot are all about Adelie. Nothing says Gentoo. Rendering without `-P` produces the
Gentoo report. If you completed the optional task, the heading also names the selected species.
:::

::: {.callout-tip collapse="true"}
## Hint
- Parameters **must have defaults**: `params$species` has to resolve when no `-P` is passed.
- `-P` takes `name:value` (`-P species:Adelie`). Pass several with repeated `-P` flags.
- The plot must use the filtered object `one`, not the full `penguins` data.
:::

**The complete, runnable solution is `solutions/day1/penguins-by-species.qmd`.** Open it to compare
your work or render it with different `-P species:` values. You can also take the source:

<a class="btn btn-primary btn-sm" href="https://raw.githubusercontent.com/cderv/raukr-2026-quarto-exercises/main/solutions/day1/penguins-by-species.qmd" role="button" target="_blank">{{< fa download >}} penguins-by-species.qmd</a>

## Troubleshooting

::: {.callout-tip collapse="true" icon=false}
## {{< fa wrench >}} If rendering fails or the output looks wrong
- **YAML is indentation-sensitive.** If rendering fails *before* any code runs, check the YAML
  header for stray spaces, tabs, or unmatched quotes.
- **Missing package?** `install.packages("<name>")` (or `renv::restore()`) and render again.
- **A figure doesn't appear?** Check the path is relative to the `.qmd` and the filename matches
  exactly (case included).
- **A cross-reference shows `?@…`?** The cell label must start with `fig-`/`tbl-`/`eq-` and be
  unique, and the `@ref` must match it.
- **A citation shows `?@gorman2014` or `[?]`?** The `@key` doesn't match a key in `references.bib`,
  or `bibliography:` isn't set.
- **The PDF's References section is empty and the entries are at the end under "Bibliography"?**
  This is expected with Typst. It builds the bibliography itself and places it last, so the
  `::: {#refs}` div only moves the list in HTML.
- **`references.bib` not found, although the download has it?** Your `.qmd` is probably not in
  `day1-intro/` (saved to the top folder, for example). Move it next to `references.bib` /
  `apa.csl` / `_brand.yml` and render again.
- **Do I need to install Typst?** No. Typst is bundled in Quarto ≥ 1.9, so there's nothing extra to
  install (no LaTeX either). Check `quarto --version`.
- **Typst PDF looks unbranded / wrong fonts?** An unbranded render succeeds silently, so check the
  **title and headings**, not the log. A branded PDF uses **Albert Sans**. An unbranded PDF uses
  Typst's default serif. Table 1 stays serif either way, so don't judge by the table. Confirm
  `_brand.yml` is in your `day1-intro/` folder (next to the `.qmd`) and you have network on the first
  render (Quarto fetches the Google fonts once). Missing the file? Download
  [`_brand.yml`](https://raw.githubusercontent.com/cderv/raukr-2026-quarto-exercises/main/day1-intro/_brand.yml)
  into the folder holding your `.qmd`.
- **Typst prints `unknown font family` warnings?** Harmless: a `gt` table names a generic
  `sans-serif` in its font stack, and Typst doesn't recognize that as a family. The PDF still
  builds, branding included.
- **Prefer to stay in HTML first?** Render HTML, get it right, *then* switch to `format: typst`.
  The content is the same.
:::

## Session {.appendix .unnumbered}

<details>
<summary>Session info</summary>

```{r}
#| label: session
sessionInfo()
```

</details>
