---
title: "Introduction to Quarto — Lab"
subtitle: "RaukR 2026 · Day 1"
author: ""
date: ""
format:
  html:
    toc-location: left   # free the right margin for the "Quarto docs" pointers
---

::: {.callout-note}
## Scope
Build one penguins document with the authoring extras, then **cite it** and render it as a
**branded Typst PDF**. Two challenges, one document: each is its own hands-on part, with a break
between them. This is your **manuscript / team-project path**, the same route from
an analysis to a submittable write-up. Everything runs on **base-R `datasets::penguins`** (R ≥ 4.5),
so there's no data to download.

**Before you start:** work through the **[Setup page](../../setup.qmd)** if you haven't. Its
`00-check-setup.R` confirms R, Quarto, and the packages this lab uses.
:::

::: {.callout-tip appearance="simple"}
You are working in the **`day1-intro/`** folder from the exercises download (open `day1-intro.Rproj`).
See the [Setup page](../../setup.qmd#get-the-materials). Everything you render is written next to
your files.
:::

Start every document with this setup (the challenges below assume it):

```{r}
#| label: setup
#| message: false
library(dplyr)
library(ggplot2)
library(gt)
library(ggokabeito)   # color-blind-safe (Okabe-Ito) scale -- see the Accessibility box

data(penguins)
penguins <- penguins |> filter(!is.na(bill_len), !is.na(bill_dep))
```

::: {.callout-tip collapse="true"}
## Coming from R Markdown? (optional, 2-minute aside)
You do **not** need to convert anything to work natively in Quarto. Start with `.qmd` directly.
But if you have an `.Rmd` you want to bring over:

- it renders **as-is** (Quarto reads `.Rmd` too).
- cell options move from the chunk header to the `#|` "hash-pipe" (one YAML option per line).
- `knitr::convert_chunk_header("doc.Rmd", output = "doc.qmd")` does the mechanical part.

That's the whole migration. The rest of this lab is native `.qmd`.
:::

## Authoring Challenge

> **Goal:** a penguins report with a cross-referenced **figure**, a cross-referenced **table**, and a
> **margin** element, rendered to HTML.

:::: {.callout-note icon=false}
## {{< fa clipboard-list >}} Tasks

::: {.column-margin}
{{< fa book-open >}} **Quarto docs**

[Cross-references](https://quarto.org/docs/authoring/cross-references.html) · [Article layout](https://quarto.org/docs/authoring/article-layout.html)
:::

1. Create `my-report.qmd` **inside your open `day1-intro/` folder** (next to `references.bib` /
   `apa.csl`, so Part 2's paths and branding resolve later) with `format: html` and the setup cell
   at the top of this page.
2. Add a scatter plot of `bill_len` versus `bill_dep` colored by `species`. Give its cell
   `#| label: fig-bill` and a `#| fig-cap:`, then refer to it with `@fig-bill` **in the prose,
   outside the cell**.
3. Add a summary table. The `gt` plumbing is below. Your job is the **cross-reference mechanic**:
   give the cell `#| label: tbl-summary` and a `#| tbl-cap:`, then refer to it with `@tbl-summary`.
   ```r
   penguins |>
     summarise(bill_len = mean(bill_len), bill_dep = mean(bill_dep), .by = species) |>
     gt()
   ```
4. Put the species counts in the **margin** with the cell option `#| column: margin`.
5. Make the report accessible: add `#| fig-alt:` to the figure and keep both color and shape mapped
   to `species`. Add `scale_color_okabe_ito()` for a color-blind-safe palette.
6. *(stretch)* Add the bill-shape ratio as a display equation, referenced with `@eq-ratio`. **Escape
   the underscores**: inside `$$…$$`, `bill_len` may render as a subscript (depending on the math
   renderer), so write:
   ```markdown
   $$
   \text{ratio} = \frac{\text{bill\_len}}{\text{bill\_dep}}
   $$ {#eq-ratio}
   ```
7. Render to HTML (`quarto render my-report.qmd`, or the editor's Render/Preview).
::::

::: {.callout-tip icon=false}
## {{< fa universal-access >}} Optional accessibility check
Your figure already has alt text and redundant color/shape encoding. To check the whole page:

- add `axe: {output: document}` under `format: html`, re-preview, and a WCAG report appears on the
  page (drop it again before you publish). See
  [Quarto's accessibility docs](https://quarto.org/docs/output-formats/html-accessibility.html).
:::

::: {.callout-note appearance="simple" icon=false}
## {{< fa circle-check >}} You should see
A numbered **Figure 1** (the bill scatter, three separate clusters), a numbered **Table 1**, and the
counts in the right margin. `@fig-bill` and `@tbl-summary` render as live links. If you completed the
stretch task, you also see a numbered **Equation 1**. Your finished figure looks like this:
:::

```{r}
#| label: fig-target
#| echo: false
#| fig-width: 6
#| fig-height: 3.4
#| fig-cap: "Target: bill length versus depth by species."
#| fig-alt: >-
#|   Scatter plot of bill depth against bill length for three penguin species in distinct colors;
#|   the species form largely separate clusters.
ggplot(penguins, aes(bill_len, bill_dep, color = species, shape = species)) +
  geom_point(alpha = 0.8) +
  scale_color_okabe_ito() +
  labs(x = "Bill length (mm)", y = "Bill depth (mm)") +
  theme_minimal(base_size = 12)
```

::: {.callout-tip collapse="true"}
## Hint
- A cell becomes a cross-referenceable figure/table only when its label **starts with**
  `fig-` / `tbl-` / `eq-` and is **unique** in the document.
- `column: margin` is a **cell option** (`#|`) for a cell, or a `::: {.column-margin}` div for
  prose.
- Display math with a label `$$ … $$ {#eq-ratio}`, referenced with `@eq-ratio`.
:::

**Need to continue with Part 2?** `starter.qmd` is a completed Part-1 document. It is already in
`day1-intro/`. Open it to continue with citations, or take the source:

<a class="btn btn-primary btn-sm" href="https://raw.githubusercontent.com/cderv/raukr-2026-quarto-exercises/main/day1-intro/starter.qmd" role="button" target="_blank">{{< fa download >}} starter.qmd</a>

The cross-references and the equation live in the **body Markdown**, outside any cell:

``` markdown
@fig-bill shows bill shape by species. @tbl-summary gives the per-species means.

$$
\text{ratio} = \frac{\text{bill\_len}}{\text{bill\_dep}}
$$ {#eq-ratio}
```

**`solutions/day1/penguins-report.qmd` is the complete Day-1 reference.** The document goes beyond
the tasks (an additional margin figure, a `glimpse`, and section labels) and also adds the
citations from the next challenge. It's in your exercises `solutions/day1/` folder. Try the challenge
first, then open it to compare, or take the source:

<a class="btn btn-primary btn-sm" href="https://raw.githubusercontent.com/cderv/raukr-2026-quarto-exercises/main/solutions/day1/penguins-report.qmd" role="button" target="_blank">{{< fa download >}} penguins-report.qmd</a>

## Citations Challenge

> **Goal:** cite the report, then render the *same document* as a **branded Typst PDF**.

::: {.callout-note}
## Starting point
If you finished the Authoring Challenge, keep going on your own document. Otherwise open
**`day1-intro/starter.qmd`** in your editor (a complete Part-1 report) and add the citations to
that.
:::

:::: {.callout-note icon=false}
## {{< fa clipboard-list >}} Tasks

::: {.column-margin}
{{< fa book-open >}} **Quarto docs**

[Citations](https://quarto.org/docs/authoring/citations.html) · [Typst](https://quarto.org/docs/output-formats/typst.html) · [Brand](https://quarto.org/docs/authoring/brand.html)
:::

1. Point the document header at the shared bibliography and citation style:
   ```yaml
   bibliography: references.bib
   csl: apa.csl
   ```
2. Cite the data source in prose with `@gorman2014`. Find (or add) the sentence ending
   *"…collected at Palmer Station, Antarctica."* and insert the citation before the period:
   `Antarctica [@gorman2014].` (don't paste a second copy).
3. Add a **References** section where the list should appear:
   ```markdown
   ## References {.unnumbered}

   ::: {#refs}
   :::
   ```
4. Give it a real **title block** so the PDF reads as a paper. **Replace** your document's
   `author:` line with:
   ```yaml
   author:
     - name: Your Name
       affiliation: Your Lab, Your University
   ```
5. Render to HTML and confirm the citation resolves to *(Gorman et al., 2014)* with a formatted
   reference list.
6. Switch the output to **Typst** and render a branded PDF. You may see `unknown font family`
   warnings on the first render. See Troubleshooting if the final PDF still uses a serif font. Two routes:
   - **CLI** — leave the header as-is and force the target:
     ```bash
     quarto render my-report.qmd --to typst
     ```
   - **Editor Render button** — set `format: typst` in the YAML header first, *then* click Render.
     The button renders whatever `format:` is declared, so while the header still says `format: html`
     the button gives you HTML, **not** a PDF.

   If you used the completed Part-1 file, render `starter.qmd` instead. The PDF is written **next to its source**
   in `day1-intro/`.
::::

::: {.callout-note appearance="simple" icon=false}
## {{< fa circle-check >}} You should see
In-text **(Gorman et al., 2014)**, a **References** section formatted in APA, and, in the Typst
PDF, the same citation and reference list, typeset with the RaukR palette. No `?@gorman2014` or
`[?]` markers anywhere.

**Check the branding applied.** An unbranded PDF builds without any error, so check the font
instead: a branded PDF is set in **Albert Sans**, an unbranded one in Typst's default serif. Serif
headings mean `_brand.yml` isn't next to your `.qmd` (see Troubleshooting).
:::

::: {.callout-tip collapse="true"}
## Hint
- The `@key` in `[@gorman2014]` must match a key in `references.bib` **exactly**.
- `csl:` sets the style (here APA). Swap the file to restyle every citation at once.
- Typst reads the `.bib` itself, so keep bibliography fields plain (avoid LaTeX macros like
  `\emph{}` in titles).
- On the first Typst render, Quarto downloads the brand's Google fonts. Give it a moment and a
  network connection.
:::

:::: {.callout-note collapse="true"}
## Solution: the header and the cite

In the YAML header:

``` yaml
bibliography: references.bib
csl: apa.csl
```

In prose:

``` markdown
The data come from Palmer Station, Antarctica [@gorman2014].
```

A References section (the list renders inside the `#refs` div):

``` markdown
## References {.unnumbered}

::: {#refs}
:::
```

For the branded PDF (`day1-intro/_brand.yml` styles it), either force the target from the CLI:

``` bash
quarto render my-report.qmd --to typst
```

or set `format: typst` in the header and use the editor's Render button.
::::

The complete cited HTML is `solutions/day1/penguins-report.qmd`. The branded Typst PDF (with R-side
`theme_brand_*` styling) is `day1-intro/sample-typst.qmd` (render it to see the final branded PDF).

## Bonus — one report per species (optional) {#sec-parameters}

> **Goal:** turn a report into a *parameterized* one: a single source that renders a **separate
> report per species** by passing the species in at render time. This is the "one QC report per
> sample / one summary per cohort" pattern. Fully optional, so skip it if you're short on time.

A **parameter** is a value declared in the YAML that your code can read as `params$…`. Change the
value at render time and the *same document* produces a different report. No copy-paste.

::: {.callout-note}
## Starting point
Open **`day1-intro/parameters-starter.qmd`**. It contains the report structure, data, and plot. Your
task is to make the species an input to the document. A complete reference is available at
**`solutions/day1/penguins-by-species.qmd`**.
:::

:::: {.callout-note icon=false}
## {{< fa clipboard-list >}} Tasks

::: {.column-margin}
{{< fa book-open >}} **Quarto docs**

[Parameters](https://quarto.org/docs/computations/parameters.html)
:::

1. Declare a parameter with a default in the header:
   ```yaml
   params:
     species: Gentoo
   ```
2. **Use it to filter** in the setup cell. `params$species` is a value. Keep the existing missing-value
   checks and add the species condition:
   ```r
   one <- penguins |> filter(species == params$species, !is.na(bill_len), !is.na(bill_dep))
   ```
3. **Use the parameter in the report.** Replace the placeholder species in the sentence with inline
   code. Make the caption and alt text track the parameter with `!expr`. Cell options are YAML, so
   the `!expr` tag tells Quarto to evaluate the R expression:
   ```yaml
   #| fig-cap: !expr paste0("Bill length versus depth for ", params$species, " penguins.")
   #| fig-alt: !expr paste0("Scatter plot of bill depth against bill length for ", params$species, " penguins.")
   ```
4. **Render a different species from the CLI** with `-P name:value`. Run this from inside
   `day1-intro/`:
   ```bash
   quarto render parameters-starter.qmd -P species:Adelie
   ```
   With no `-P`, the default (`Gentoo`) is used.
::::

::: {.callout-tip collapse="true"}
## Stretch: dynamic headings and standalone files
Build the section heading from the parameter with a cell that emits Markdown:

````markdown
```{{r}}
#| echo: false
#| output: asis
cat("##", params$species, "at a glance\n")
```
````

To keep two named reports, render each as a self-contained HTML file:

```bash
quarto render parameters-starter.qmd -P species:Adelie    --output adelie.html    --embed-resources
quarto render parameters-starter.qmd -P species:Chinstrap --output chinstrap.html --embed-resources
```

`--output` prevents the second render from overwriting the first. `--embed-resources` puts each
plot inside its HTML file, so the reports do not share a figure-support folder.
:::

::: {.callout-note appearance="simple" icon=false}
## {{< fa circle-check >}} You should see
Rendering with `-P species:Adelie` gives a report whose sentence, figure caption, alt text, point
count, and plot are all about Adelie. Nothing says Gentoo. Rendering without `-P` produces the
Gentoo report. If you completed the stretch task, the heading also names the selected species.
:::

::: {.callout-tip collapse="true"}
## Hint
- Parameters **must have defaults**: `params$species` has to resolve when no `-P` is passed.
- `-P` takes `name:value` (`-P species:Adelie`). Pass several with repeated `-P` flags.
- The plot must use the filtered object `one`, not the full `penguins` data.
:::

**The complete, runnable solution is `solutions/day1/penguins-by-species.qmd`.** Open it to compare
your work or render it with different `-P species:` values. You can also take the source:

<a class="btn btn-primary btn-sm" href="https://raw.githubusercontent.com/cderv/raukr-2026-quarto-exercises/main/solutions/day1/penguins-by-species.qmd" role="button" target="_blank">{{< fa download >}} penguins-by-species.qmd</a>

## Troubleshooting

::: {.callout-tip collapse="true" icon=false}
## {{< fa wrench >}} If rendering fails or the output looks wrong
- **YAML is indentation-sensitive.** If rendering fails *before* any code runs, check the YAML
  header for stray spaces, tabs, or unmatched quotes.
- **Missing package?** `install.packages("<name>")` (or `renv::restore()`) and render again.
- **A figure doesn't appear?** Check the path is relative to the `.qmd` and the filename matches
  exactly (case included).
- **A cross-reference shows `?@…`?** The cell label must start with `fig-`/`tbl-`/`eq-` and be
  unique, and the `@ref` must match it.
- **A citation shows `?@gorman2014` or `[?]`?** The `@key` doesn't match a key in `references.bib`,
  or `bibliography:` isn't set.
- **Do I need to install Typst?** No. Typst is bundled in Quarto ≥ 1.9, so there's nothing extra to
  install (no LaTeX either). Check `quarto --version`.
- **Typst PDF looks unbranded / wrong fonts?** An unbranded render succeeds silently, so check the
  font, not the log: branded is **Albert Sans**, unbranded is Typst's default serif. Confirm
  `_brand.yml` is in your `day1-intro/` folder (next to the `.qmd`) and you have network on the
  first render (Quarto fetches the Google fonts once).
- **Typst prints `unknown font family` warnings?** Harmless: a `gt` table names a generic
  `sans-serif` in its font stack, and Typst doesn't recognize that as a family. The PDF still
  builds, branding included.
- **Prefer to stay in HTML first?** Render HTML, get it right, *then* switch to `format: typst`.
  The content is the same.
:::

## Session {.appendix .unnumbered}

<details>
<summary>Session info</summary>

```{r}
#| label: session
sessionInfo()
```

</details>
