---
title: "Introduction to Quarto — Lab"
subtitle: "RaukR 2026 · Day 1"
author: ""
date: ""
---

::: {.callout-note}
## Scope
Build one penguins document with the authoring extras, then **cite it** and render it as a
**branded Typst PDF**. Two challenges, one document — each is its own hands-on part, with the
between-parts break in between. This is your **manuscript / team-project path** — the same route from
an analysis to a submittable write-up. Everything runs on **base-R `datasets::penguins`** (R ≥ 4.5)
— no data to download.

**Packages:** `dplyr`, `ggplot2`, `gt`, `ggokabeito` (a colour-blind-safe palette — see the
Accessibility box below), plus the `knitr`/`rmarkdown` engine every `.qmd` with R code needs. If
you've done the **[Setup page](../../setup.qmd)** (with its one-command `quarto check`), you're ready;
otherwise install them once from R:
`install.packages(c("dplyr", "ggplot2", "gt", "ggokabeito"))`. The citation payoff also uses the
shared `references.bib` and `apa.csl` shipped in this folder.
:::

Start every document with this setup — the challenges below assume it:

```{r}
#| label: setup
#| message: false
library(dplyr)
library(ggplot2)
library(gt)
library(ggokabeito)   # colour-blind-safe (Okabe-Ito) scale -- see the Accessibility box

data(penguins)
penguins <- penguins |> filter(!is.na(bill_len), !is.na(bill_dep))
```

::: {.callout-tip collapse="true"}
## Coming from R Markdown? (2-minute aside — optional)
You do **not** need to convert anything to work natively in Quarto — start with `.qmd` directly.
But if you have an `.Rmd` you want to bring over:

- it renders **as-is** — Quarto reads `.Rmd` too;
- cell options move from the chunk header to the `#|` "hash-pipe" (one YAML option per line);
- `knitr::convert_chunk_header("doc.Rmd", output = "doc.qmd")` does the mechanical part.

That's the whole migration. The rest of this lab is native `.qmd`.
:::

## Authoring Challenge

> **Goal:** a penguins report with a cross-referenced **figure**, a cross-referenced **table**, a
> **margin** element, and a display **equation** — rendered to HTML.

::: {.callout-note icon=false}
## {{< fa clipboard-list >}} Tasks
1. Create a new `.qmd` **inside `labs/quarto/`** in your cloned repo (next to `references.bib` /
   `apa.csl`, so Part 2's paths and branding resolve later) with `format: html` and the setup cell
   at the top of this page.
2. Add a scatter plot of `bill_len` versus `bill_dep` colored by `species`. Give its cell
   `#| label: fig-bill` and a `#| fig-cap:`, then — **in the prose, outside the cell** — refer to
   it with `@fig-bill`.
3. Add a summary table. The `gt` plumbing is below; your job is the **cross-reference mechanic** —
   give the cell `#| label: tbl-summary` and a `#| tbl-cap:`, then refer to it with `@tbl-summary`.
   ```r
   penguins |>
     summarise(bill_len = mean(bill_len), bill_dep = mean(bill_dep), .by = species) |>
     gt()
   ```
4. Put the species counts in the **margin** with the cell option `#| column: margin`.
5. *(stretch)* Add the bill-shape ratio as a display equation, referenced with `@eq-ratio`. **Escape
   the underscores** — inside `$$…$$`, `bill_len` may render as a subscript (depending on the math
   renderer), so write:
   ```markdown
   $$
   \text{ratio} = \frac{\text{bill\_len}}{\text{bill\_dep}}
   $$ {#eq-ratio}
   ```
6. Render to HTML (`quarto render your-doc.qmd`, or the editor's Render/Preview).
:::

::: {.callout-tip icon=false}
## {{< fa universal-access >}} Make it accessible
The worked report does all three — do the same in yours:

- add a `#| fig-alt:` to every figure;
- add `scale_color_okabe_ito()` (from `ggokabeito`) to your plot, keeping `shape = species`;
- add `axe: {output: document}` under `format: html`, re-preview, and a WCAG report appears on the
  page (drop it again before you publish). See
  [Quarto's accessibility docs](https://quarto.org/docs/output-formats/html-accessibility.html).
:::

::: {.callout-note appearance="simple" icon=false}
## {{< fa circle-check >}} You should see
A numbered **Figure 1** (the bill scatter, three separate clusters), a numbered **Table 1**, the
counts sitting in the right margin, and an **Equation 1** — with `@fig-bill` / `@tbl-summary` /
`@eq-ratio` rendering as live "Figure 1" / "Table 1" / "Equation 1" links. Your finished figure
looks like this:
:::

```{r}
#| label: fig-target
#| echo: false
#| fig-width: 6
#| fig-height: 3.4
#| fig-cap: "Target: bill length versus depth by species."
#| fig-alt: >-
#|   Scatter plot of bill depth against bill length for three penguin species in distinct colors;
#|   the species form largely separate clusters.
ggplot(penguins, aes(bill_len, bill_dep, color = species, shape = species)) +
  geom_point(alpha = 0.8) +
  scale_color_okabe_ito() +
  labs(x = "Bill length (mm)", y = "Bill depth (mm)") +
  theme_minimal(base_size = 12)
```

::: {.callout-tip collapse="true"}
## Hint
- A cell becomes a cross-referenceable figure/table only when its label **starts with**
  `fig-` / `tbl-` / `eq-` and is **unique** in the document.
- `column: margin` is a **cell option** (`#|`) for a cell, or a `::: {.column-margin}` div for
  prose.
- Display math with a label: `$$ … $$ {#eq-ratio}` — reference it with `@eq-ratio`.
:::

**The complete, runnable solution is `starter.qmd`** — a known-good Part-1 report with all these cells.
Open it from your cloned `labs/quarto/` to compare, or take the source:

<!-- Raw-GitHub download button -- RaukR house pattern (see project-context.md; no `downloadthis`).
     The URL hard-codes owner/repo/branch: it goes live once this branch merges to `main`, and must
     switch to `NBISweden/raukr-2026/...` when the material folds into the NBIS course site. -->
<a class="btn btn-primary btn-sm" href="https://raw.githubusercontent.com/cderv/raukr-2026-quarto/refs/heads/main/labs/quarto/starter.qmd" role="button" target="_blank">{{< fa download >}} starter.qmd</a>

The cross-references and the equation live in the **body Markdown**, outside any cell:

``` markdown
@fig-bill shows bill shape by species; @tbl-summary gives the per-species means.

$$
\text{ratio} = \frac{\text{bill\_len}}{\text{bill\_dep}}
$$ {#eq-ratio}
```

**[`penguins-report.qmd`](penguins-report.qmd) is a reference, not a checklist.** The complete worked
document goes beyond the tasks (a margin culmen figure, a `glimpse`, section labels) and also adds the
citations from the next challenge. Prefer to open the whole thing in your editor? Take the source:

<!-- Raw-GitHub download button -- RaukR house pattern (see project-context.md; no `downloadthis`
     extension). The URL hard-codes owner/repo/branch: it goes live once this branch merges to `main`,
     and must switch to `NBISweden/raukr-2026/...` when the material folds into the NBIS course site. -->
<a class="btn btn-primary btn-sm" href="https://raw.githubusercontent.com/cderv/raukr-2026-quarto/refs/heads/main/labs/quarto/penguins-report.qmd" role="button" target="_blank">{{< fa download >}} penguins-report.qmd</a>

## Citations Challenge

> **Goal:** cite the report, then render the *same document* as a **branded Typst PDF** — the
> manuscript payoff, no LaTeX.

::: {.callout-note}
## Starting point
If you finished the Authoring Challenge, keep going on your own document. Otherwise open
**`labs/quarto/starter.qmd`** in your editor — a known-good Part-1 report — and add the citations to
that. Nobody is stranded by the break.
:::

::: {.callout-note icon=false}
## {{< fa clipboard-list >}} Tasks
1. Point the document header at the shared bibliography and citation style:
   ```yaml
   bibliography: references.bib
   csl: apa.csl
   ```
2. Cite the data source in prose with `@gorman2014`. Find (or add) the sentence ending
   *"…collected at Palmer Station, Antarctica."* and insert the citation before the period:
   `Antarctica [@gorman2014].` (don't paste a second copy).
3. Add a **References** section where the list should appear:
   ```markdown
   ## References {.unnumbered}

   ::: {#refs}
   :::
   ```
4. Give it a real **title block** so the PDF reads as a paper — **replace** your document's
   `author:` line with:
   ```yaml
   author:
     - name: Your Name
       affiliation: Your Lab, Your University
   ```
5. Render to HTML and confirm the citation resolves to *(Gorman et al., 2014)* with a formatted
   reference list.
6. Switch the output to **Typst** and render a branded PDF. *(Heads-up: you'll see a stack of
   `unknown font family` warnings as it renders — they're harmless; the Troubleshooting box below
   explains them.)* Two routes:
   - **CLI** — leave the header as-is and force the target:
     ```bash
     quarto render your-doc.qmd --to typst
     ```
   - **Editor Render button** — set `format: typst` in the YAML header first, *then* click Render.
     The button renders whatever `format:` is declared, so while the header still says `format: html`
     the button gives you HTML, **not** a PDF.

   Where the PDF lands: rendering the shipped `starter.qmd` writes the PDF under `_site/…` (it's in
   the site's render list); a brand-new doc renders the PDF next to its source.
:::

::: {.callout-note appearance="simple" icon=false}
## {{< fa circle-check >}} You should see
In-text **(Gorman et al., 2014)**, a **References** section formatted in APA, and — in the Typst
PDF — the same citation and reference list, typeset with the RaukR palette. No `?@gorman2014` or
`[?]` markers anywhere.
:::

::: {.callout-tip collapse="true"}
## Hint
- The `@key` in `[@gorman2014]` must match a key in `references.bib` **exactly**.
- `csl:` sets the style (here APA). Swap the file to restyle every citation at once.
- Typst reads the `.bib` itself, so keep bibliography fields plain — avoid LaTeX macros like
  `\emph{}` in titles.
- On the first Typst render, Quarto downloads the brand's Google fonts; give it a moment and a
  network connection.
:::

```{r}
#| label: sol-citations
#| code-fold: true
#| code-summary: "Solution — the header and the cite"
#| eval: false
# In the YAML header:
#   bibliography: references.bib
#   csl: apa.csl
#
# In prose:
#   The data come from Palmer Station, Antarctica [@gorman2014].
#
# A References section (the list renders inside the #refs div):
#   ## References {.unnumbered}
#   ::: {#refs}
#   :::
#
# For the branded PDF (the project _brand.yml styles it), either force the target from the CLI:
#   quarto render your-doc.qmd --to typst
# or set `format: typst` in the header and use the editor's Render button.
```

The complete cited HTML is [`penguins-report.qmd`](penguins-report.qmd); the branded Typst PDF —
with R-side `theme_brand_*` styling — is [`sample-typst.qmd`](sample-typst.qmd) (the link opens the
rendered PDF, not a web page).

## Bonus — one report per species (optional) {#sec-parameters}

> **Goal:** turn a report into a *parameterized* one — a single source that renders a **separate
> report per species** by passing the species in at render time. This is the "one QC report per
> sample / one summary per cohort" pattern. Fully optional, self-service — skip it if you're out of
> time.

A **parameter** is a value declared in the YAML that your code can read as `params$…`. Change the
value at render time and the *same document* produces a different report — no copy-paste.

::: {.callout-note icon=false}
## {{< fa clipboard-list >}} Tasks
1. Declare a parameter with a default in the header:
   ```yaml
   params:
     species: Gentoo
   ```
2. **Use it to filter** in your setup cell — `params$species` is just a value:
   ```r
   one <- penguins |> filter(species == params$species)
   ```
3. **Build a heading from the parameter** with `#| output: asis` (it writes the cell's text output
   as raw Markdown, so `cat()` becomes a real heading):
   ````markdown
   ```{{r}}
   #| echo: false
   #| output: asis
   cat("## ", params$species, "at a glance\n")
   ```
   ````
4. **Make the caption track the parameter** with `!expr` (evaluate R in a cell option):
   ```yaml
   #| fig-cap: !expr paste0("Bill length versus depth for ", params$species, " penguins.")
   ```
5. **Render a different species from the CLI** — same `quarto render` you used for the Typst PDF,
   with `-P name:value` (no Render button for this; the parameter override is a command-line thing):
   ```bash
   quarto render penguins-by-species.qmd -P species:Adelie
   quarto render penguins-by-species.qmd -P species:Chinstrap
   ```
   With no `-P`, the default (`Gentoo`) is used.
:::

::: {.callout-note appearance="simple" icon=false}
## {{< fa circle-check >}} You should see
Rendering with `-P species:Adelie` gives a report whose title sentence, section heading, figure
caption, and point count are **all** about Adelie — nothing says Gentoo. Change the one word after
`species:` and the whole report follows.
:::

::: {.callout-tip collapse="true"}
## Hint
- Parameters **must have defaults** — `params$species` has to resolve when no `-P` is passed.
- `-P` takes `name:value` (`-P species:Adelie`). Pass several with repeated `-P` flags.
- `#| output: asis` is what lets code *emit Markdown* (a heading here); without it, `cat()` output
  would show as a printed code result, not a heading.
:::

**The complete, runnable solution is [`penguins-by-species.qmd`](penguins-by-species.qmd)** — open it
from your cloned `labs/quarto/` and render it with different `-P species:` values, or take the source:

<!-- Raw-GitHub download button -- RaukR house pattern (see project-context.md; no `downloadthis`).
     The URL hard-codes owner/repo/branch: it goes live once this branch merges to `main`, and must
     switch to `NBISweden/raukr-2026/...` when the material folds into the NBIS course site. -->
<a class="btn btn-primary btn-sm" href="https://raw.githubusercontent.com/cderv/raukr-2026-quarto/refs/heads/main/labs/quarto/penguins-by-species.qmd" role="button" target="_blank">{{< fa download >}} penguins-by-species.qmd</a>

## Troubleshooting

::: {.callout-tip collapse="true" icon=false}
## {{< fa wrench >}} If a render misbehaves
- **YAML is indentation-sensitive.** If rendering fails *before* any code runs, check the YAML
  header for stray spaces, tabs, or unmatched quotes.
- **Missing package?** `install.packages("<name>")` (or `renv::restore()`) and render again.
- **A figure doesn't appear?** Check the path is relative to the `.qmd` and the filename matches
  exactly (case included).
- **A cross-reference shows `?@…`?** The cell label must start with `fig-`/`tbl-`/`eq-` and be
  unique, and the `@ref` must match it.
- **A citation shows `?@gorman2014` or `[?]`?** The `@key` doesn't match a key in `references.bib`,
  or `bibliography:` isn't set.
- **Do I need to install Typst?** No — Typst is bundled in Quarto ≥ 1.9, so there's nothing extra to
  install (no LaTeX either); check `quarto --version`.
- **Typst PDF looks unbranded / wrong fonts?** Confirm `_brand.yml` is at the project root and you
  have network on the first render (Quarto fetches the Google fonts once).
- **Typst prints `unknown font family` warnings?** Harmless — a plain `gt` table emits them for its
  default font stack; the PDF still builds. (The fully brand-styled table via `theme_brand_*` is in
  `sample-typst.qmd`.)
- **Prefer to stay in HTML first?** Render HTML, get it right, *then* switch to `format: typst` —
  the content is the same.
:::

## Session {.appendix .unnumbered}

<details>
<summary>Session info</summary>

```{r}
#| label: session
sessionInfo()
```

</details>
