---
title: "Bill dimensions across Antarctic penguin species"
subtitle: "The running Day-1 document — authoring value-adds, then citations"
format: html
bibliography: references.bib
csl: apa.csl
---

::: {.callout-note}
## About this document
This is the **shared running document** for Day 1: a small penguins report that carries the
authoring value-adds (figures, cross-references, a `gt` table, margin layout, math) and then the
citations. The Day-1 lab grows a version of it, then renders it as a branded PDF
with **Typst** (a modern typesetting system that ships inside Quarto, so there is no LaTeX to
install). Dataset: base-R `datasets::penguins` (R ≥ 4.5), columns `bill_len`, `bill_dep`,
`flipper_len`, `body_mass`.
:::

```{r}
#| label: setup
#| message: false
library(dplyr)
library(ggplot2)
library(gt)
library(ggokabeito)

data(penguins)
penguins <- penguins |> filter(!is.na(bill_len), !is.na(bill_dep))
```

## The data {#sec-data}

We use the `penguins` dataset: body measurements for `{r} nrow(penguins)` penguins of
`{r} length(unique(penguins$species))` species (`{r} knitr::combine_words(levels(penguins$species))`),
collected at Palmer Station, Antarctica [@gorman2014]. The data reached R first through the
**palmerpenguins** package [@horst2020] and now ships in base R's `datasets`.

```{r}
#| label: glimpse
glimpse(penguins)
```

The species are not evenly represented:

```{r}
#| label: counts
#| column: margin
penguins |>
  count(species, name = "n") |>
  knitr::kable()
```

## Bill length and depth {#sec-bill}

Each penguin's bill is summarized by two measurements: **length** and **depth** (the *culmen*,
@fig-culmen). Depth is the vertical thickness at the base. Length runs from base to tip.

```{r}
#| label: fig-culmen
#| echo: false
#| column: margin
#| fig-align: center
#| fig-alt: >-
#|   Line drawing of a penguin head showing culmen length measured along the top of the bill
#|   from base to tip, and culmen depth measured as the vertical thickness of the bill at its base.
#| fig-cap: "Culmen length and depth. Artwork by Allison Horst [@horst2020]."
knitr::include_graphics(
  "https://allisonhorst.github.io/palmerpenguins/reference/figures/culmen_depth.png"
)
```

@fig-bill shows the two measurements together: within a species the bill has a characteristic
shape, so the species separate cleanly.

```{r}
#| label: fig-bill
#| warning: false
#| fig-width: 5
#| fig-cap: "Bill length versus depth, colored by species."
#| fig-cap-location: margin
#| fig-alt: >-
#|   Scatter plot of bill depth against bill length for penguins of three species, each drawn in a
#|   distinct color and shape. The three species form largely separate clusters, showing that bill
#|   shape distinguishes them.
ggplot(penguins, aes(x = bill_len, y = bill_dep, color = species, shape = species)) +
  geom_point(alpha = 0.8) +
  scale_color_okabe_ito() +   # color-blind-safe palette; shape also encodes species
  labs(x = "Bill length (mm)", y = "Bill depth (mm)", color = "Species", shape = "Species") +
  theme_minimal(base_size = 12)
```

A compact way to describe bill shape is the **length-to-depth ratio** (@eq-ratio): larger values
mean a longer, shallower bill.

$$
\text{ratio} = \frac{\text{bill\_len}}{\text{bill\_dep}}
$$ {#eq-ratio}

## A summary table {#sec-table}

@tbl-summary reports the mean bill measurements and body mass per species.

```{r}
#| label: tbl-summary
#| tbl-cap: "Mean measurements per species (values rounded)."
penguins |>
  summarise(
    n = n(),
    bill_len = mean(bill_len),
    bill_dep = mean(bill_dep),
    body_mass = mean(body_mass, na.rm = TRUE),
    .by = species
  ) |>
  gt() |>
  tab_header(title = "Penguins of the Palmer Archipelago") |>
  tab_spanner(label = md("**Bill (mm)**"), columns = c(bill_len, bill_dep)) |>
  cols_label(
    species = "Species", n = "N",
    bill_len = "Length", bill_dep = "Depth", body_mass = "Body mass (g)"
  ) |>
  fmt_number(columns = c(bill_len, bill_dep), decimals = 1) |>
  fmt_number(columns = body_mass, decimals = 0) |>
  opt_stylize(style = 6, color = "cyan")
```

## References {.unnumbered}

::: {#refs}
:::

## Session {.appendix .unnumbered}

<details>
<summary>Session info</summary>

```{r}
#| label: session
sessionInfo()
```

</details>
