---
title: "Quarto projects — Lab"
subtitle: "RaukR 2026 · Day 2"
author: ""
date: ""
format:
  html:
    toc-location: left   # free the right margin for the "Quarto docs" pointers
---

```{r}
#| label: setup
#| include: false
#| message: false
library(dplyr)
library(ggplot2)
library(ggokabeito)   # color-blind-safe (Okabe-Ito) scale, same as Day 1
data(penguins)
penguins <- penguins |> filter(!is.na(bill_len), !is.na(bill_dep), !is.na(body_mass))
```

::: {.callout-note}
## Scope
Turn a set of `.qmd` files into a navigable, **branded website**. Then use **freeze** for
reproducible builds and render the site to a publishable folder. Publishing online is optional.
Everything uses base-R `datasets::penguins` (R ≥ 4.5).

**Before you start:** work through the **[Setup page](../../setup.qmd)** if you haven't. Its
`00-check-setup.R` confirms R, Quarto, and the packages this lab uses.
:::

**Starting point.** Open your
**`day2-projects/`** folder from the exercises download (double-click `day2-projects.Rproj`, see the
[Setup page](../../setup.qmd#get-the-materials)). It holds four content pages that already render on
their own but are **not yet a project**:

- `index.qmd` — a home page
- `analysis.qmd` — a penguins analysis with a figure and a table
- `reports/methods.qmd` and `reports/data-notes.qmd` — two short supporting pages, in their own
  folder because you will treat that folder differently later on

Render one page to confirm it works: `quarto render analysis.qmd` creates `analysis.html` next to
its source. A finished reference is in **`solutions/day2/`**.

## Website Challenge

> **Goal:** turn the `day2-projects/` pages into a navigable, **branded website** that renders locally to a
> `_site/` folder.

Complete the four core steps below. The listing is a stretch task. Each step includes a checkpoint,
a hint, and a solution.

::: {.column-margin}
{{< fa book-open >}} **Quarto docs**

[Websites](https://quarto.org/docs/websites/) · [Projects](https://quarto.org/docs/projects/quarto-projects.html) · [Brand](https://quarto.org/docs/authoring/brand.html)
:::

Use **`solutions/day2/`** to compare your work or complete missed steps.

### Make it a project

**Goal.** Make Quarto treat `day2-projects/` as one project, so that a single command builds every
page in it into an output folder.

This stanza declares a project. Where it goes, and how you then build the folder, is the work:

```yaml
project:
  type: website
  output-dir: _site
```

(`_site` is already the default output folder for a website. We write it explicitly here so the
publishable folder is visible in the configuration.)

::: {.callout-note appearance="simple" icon=false}
## {{< fa circle-check >}} You should see
One command, run with **no file argument**, builds **all four** pages at once, and the HTML is written to a
new `day2-projects/_site/` folder instead of next to each `.qmd`. Before this step, rendering
`analysis.qmd` put `analysis.html` right beside it.
:::

::: {.callout-tip collapse="true"}
## Hint
**Ask yourself:**

You have folders above `day2-projects/` and files inside it. Which one does Quarto look at to decide
where a project starts, and in which direction does it search?

**Mechanism:**

Quarto walks **up** from the file being rendered and stops at the **nearest** `_quarto.yml`. That
file marks the project root. Nothing sits above `day2-projects/` in your download, so a `_quarto.yml`
placed inside it makes that folder the project and nothing else can claim your pages.

A `quarto render` with **no file argument** builds the whole project. `quarto render index.qmd`
still builds just the one page.
:::

::: {.callout-note collapse="true"}
## Solution
Create `day2-projects/_quarto.yml`:

```{.yaml filename="day2-projects/_quarto.yml"}
project:
  type: website
  output-dir: _site
```

Then, from inside `day2-projects/`:

```bash
quarto render
```
:::

### Build the navigation

**Goal.** Give the site a title and a top navigation bar, so a reader can move between the home and
analysis pages without editing the URL.

Add `index.qmd` (Home) and `analysis.qmd` (Analysis) to the navigation. Give the site the title
**Penguin Lab**, and use the ready-made **cosmo** theme so it does not look like raw HTML.

From this step on, preview the site with `quarto preview` from inside `day2-projects/`. On Day 1 you
previewed one document. From the project root, the command serves the whole site and
re-renders a page when you save it.

::: {.callout-note appearance="simple" icon=false}
## {{< fa circle-check >}} You should see
The preview opens the home page in your browser. A bar across the top carries the site title, two
entries, **Home** and **Analysis**, and a magnifier (the built-in site search, on by default).
Clicking Analysis moves you to the other page, which carries this figure:
:::

```{r}
#| label: fig-target
#| echo: false
#| fig-width: 6
#| fig-height: 3.2
#| fig-cap: "The analysis page's figure: body mass by species."
#| fig-alt: >-
#|   Boxplot of penguin body mass in grams for Adelie, Chinstrap and Gentoo species; Gentoo is
#|   clearly the heaviest, Adelie and Chinstrap overlap.
ggplot(penguins, aes(species, body_mass, fill = species)) +
  geom_boxplot(show.legend = FALSE) +
  scale_fill_okabe_ito() +
  labs(x = NULL, y = "Body mass (g)")
```

::: {.callout-tip collapse="true"}
## Hint
The `project:` key configures the *build*. A navbar is not about building, it is about the site
itself. What would you expect that second top-level key to be called, and what does each navbar entry
need to know?

A `website:` key sits next to `project:` at the top level. It takes a `title:`, and a `navbar:` with a
`left:` (or `right:`) list. Each entry is a pair: `href:` pointing at the **source** `.qmd` (not the
built `.html`), and `text:` for the label.

The theme is a **format** option, so it goes under `format: html:`.
:::

::: {.callout-note collapse="true"}
## Solution
```{.yaml filename="day2-projects/_quarto.yml"}
project:
  type: website
  output-dir: _site

website:
  title: "Penguin Lab"
  navbar:
    left:
      - href: index.qmd
        text: Home
      - href: analysis.qmd
        text: Analysis

format:
  html:
    theme: cosmo
```

Then, from inside `day2-projects/`:

```bash
quarto preview
```

Leave it running: it re-renders and refreshes as you save. `quarto render` still works when you only
want the files.
:::

::: {.callout-tip collapse="true"}
## Try the other menu: a sidebar built for you
Swap the `navbar:` block for this, re-render, and compare:

```yaml
  sidebar:
    contents: auto
```

`contents: auto` builds the menu from the files it finds, so you never list pages by hand. It is the
right choice for a site that grows. A navbar is the right choice when you want to control exactly
what appears and in what order. Put the navbar back before you carry on, or keep both.
:::

::: {.callout-note}
## Linking pages, and what does not carry across
The Day-1 rules now meet a multi-page site.

`index.qmd` already links to the analysis with a plain Markdown link, `[analysis](analysis.qmd)`.
That is deliberate. **Across** pages there is no auto-numbering (that is a *book* feature, not a
website one), so a plain link is how you connect website pages.

**Within** `analysis.qmd`, yesterday's rule still holds: `@tbl-means` resolves to *Table 1* because
the reference and table share one page.
:::

### Brand it

**Goal.** Give the site one shared palette and typeface from a `_brand.yml`, the same file you met on
Day 1.

::: {.column-margin}
{{< fa book-open >}} **Quarto docs**

[Brand](https://quarto.org/docs/authoring/brand.html) ·
[Theme layers](https://quarto.org/docs/output-formats/html-themes-more.html)
:::

Start by creating this file and rendering the site:

```{.yaml filename="day2-projects/_brand.yml"}
color:
  palette:
    teal: "#4C979F"
  primary: teal
typography:
  fonts:
    - family: Albert Sans
      source: google
  base:
    family: Albert Sans
```

::: {.callout-note appearance="simple" icon=false}
## {{< fa circle-check >}} You should see
After adding `_brand.yml`, the **font changes** but the navbar and links remain cosmo blue. This is
expected. After correcting the theme order, the navbar and links become teal.
:::

::: {.callout-tip collapse="true"}
## Hint
Compare three parts of the rendered site: the body font, the navbar, and the links. The font changes,
but the navbar and links remain blue.

`theme:` takes a **list**, even when you write a single value. So Quarto is already combining
something with your `cosmo`. If two layers both define a primary color, which one would you expect to
end up on the page?

**Mechanism:**

`brand` is a theme **layer**, and it is the layer that reads your `_brand.yml`. A bare
`theme: cosmo` expands to `[brand, cosmo]`, and **later layers win**. Quarto applies cosmo's palette
after the brand palette, so the site uses cosmo blue.

The fonts come through whichever way round you write it, which is exactly what makes this look like
it worked.
:::

::: {.callout-note collapse="true"}
## Solution
Put `brand` **last** in the list:

```{.yaml filename="day2-projects/_quarto.yml"}
format:
  html:
    theme: [cosmo, brand]
```

Dropping `theme:` altogether works too: with no named theme to fight, the brand is all there is.
:::

### Scope options to one folder

**Goal.** Fold the code on the two pages in `reports/`, and only there. The home and analysis pages
keep showing their code as before.

You have not looked at `day2-projects/reports/` yet. It holds two short pages, `methods.qmd` and
`data-notes.qmd`, each with a small code cell.

::: {.callout-note appearance="simple" icon=false}
## {{< fa circle-check >}} You should see
On the two `reports/` pages, each code cell is collapsed behind a **Show the code** toggle that
expands when clicked. On `index.qmd` and `analysis.qmd`, nothing changes. You did not edit any of the
four `.qmd` files to get this.
:::

::: {.callout-tip collapse="true"}
## Hint
Project-wide options live in `_quarto.yml` at the project root. You want the same kind of option, but
reaching one folder only. Where would you put a file so that Quarto applies it to that folder and
nowhere else?

A `_metadata.yml` inside a folder applies to every document in that folder (and below it). Quarto
merges it on top of the project defaults, so you write only what differs.

It takes the same **document** options you would write in `_quarto.yml`, so what differs is the
reach, not the syntax. The option that folds code is `code-fold`, and `code-summary` sets the toggle
label.
:::

::: {.callout-note collapse="true"}
## Solution
Create `day2-projects/reports/_metadata.yml`:

```{.yaml filename="day2-projects/reports/_metadata.yml"}
code-fold: true
code-summary: "Show the code"
```

Re-render the project. Nothing else changes.
:::

### Index the reports *(stretch)*

*Only start this after the four core steps work.*

**Goal.** Put a table of the `reports/` pages on the home page, built by Quarto rather than typed by
you, so that adding a page there shows up without editing `index.qmd`.

Docs: [Website listings](https://quarto.org/docs/websites/website-listings.html).

::: {.callout-note appearance="simple" icon=false}
## {{< fa circle-check >}} You should see
A table on the home page with one row per page in `reports/`, showing each title and its description
(both pages already carry a `description:` in their front matter). Add a third page to `reports/`,
re-render, and it appears in the table on its own.
:::

::: {.callout-tip collapse="true"}
## Hint
This is a property of one page, not of the whole site. So where does its configuration go, and how
does the page say *which* documents to index?

A `listing:` key in the **front matter of `index.qmd`** points at a folder with `contents:`. Give it
an `id:`, then place the result where you want it on the page with an empty div carrying that id:

```markdown
::: {#reports}
:::
```

Without the `id:` and the div, the listing renders at the bottom of the page.
:::

::: {.callout-note collapse="true"}
## Solution
In the front matter of `day2-projects/index.qmd`:

```{.yaml filename="day2-projects/index.qmd"}
---
title: "Penguin Lab"
listing:
  id: reports
  contents: reports
  type: table
  fields: [title, description]
  sort: title
---
```

Then, in the body, where you want the table:

```markdown
## Reports

::: {#reports}
:::
```
:::

## Reproducible Build Challenge {#ship-it-challenge}

> **Goal:** make the project's builds **reproducible**, bring your own work onto the site, and render
> it all to a publishable `_site/` folder.

::: {.callout-note}
## Starting point
Keep going with your Website Challenge `day2-projects/`. If you didn't finish it, open the ready-made
**`solutions/day2/`** project (a complete Part-1 site) and do the steps below in there.
:::

If `quarto preview` is still running, stop it (`Ctrl` + `C` in the terminal). The freeze steps below
use renders that you start yourself. To check a page after a render, open the built file (for
example `_site/analysis.html`) directly in your browser and reload it.

**Core tasks:** prove that freeze works, then add your Day-1 report to the site. If time is short,
stop after those two. The guided `freeze: true` experiment completes the core. We will compare
cache and freeze together. The dashboard demo and publishing are optional.

::: {.column-margin}
{{< fa book-open >}} **Quarto docs**

[Project execution](https://quarto.org/docs/projects/code-execution.html) ·
[Dashboard layout](https://quarto.org/docs/dashboards/layout.html) ·
[Posit Connect Cloud](https://quarto.org/docs/publishing/posit-connect-cloud.html)
:::

### Make the build skip unchanged pages

**Goal.** Right now every `quarto render` re-runs all the R on every page, even the pages you have not
touched. Make Quarto reuse the previous results for a page whose source has not changed. Then
**prove** that it worked.

Prove that the code did not run again. Use a value that changes on every execution.

::: {.callout-note appearance="simple" icon=false}
## {{< fa circle-check >}} You should see
Render twice with no edits in between. On the second render, the R on your pages does not run again,
and you can **show** it: something on the rendered page that would certainly be different if the code
had run, and is not. Then edit the code on one page only and render again. Only that page runs.
:::

::: {.callout-tip collapse="true"}
## Hint
**Ask yourself:**

Two questions, one for each half. For the setting: this is about *executing* code across the whole
project, so which top-level key would carry it? For the proof: what could a cell print that is
**guaranteed** to differ between two runs a minute apart?

**Mechanism:**

An `execute:` key at the top level of `_quarto.yml` carries execution options for the project. The one
you want is `freeze`, and `auto` means "re-execute only when this document's own source changes".

For the proof: print the current time. If the value on the page is identical across two renders, the
code that produced it did not run.
:::

::: {.callout-note collapse="true"}
## Solution
In `day2-projects/_quarto.yml`:

```{.yaml filename="day2-projects/_quarto.yml"}
execute:
  freeze: auto
```

For the visible tell, add a cell to `analysis.qmd`:

```r
cat(format(Sys.time()))
```

Then, from inside `day2-projects/`:

```bash
quarto render     # runs the code, writes _freeze/ and _site/
quarto render     # no edits: cells are skipped, and the printed time is unchanged
```

Now change the plot code in `analysis.qmd` and render again. Only that page runs, and its timestamp
changes.
:::

::: {.callout-note}
## Why this matters beyond the wait
The results now live in `_freeze/`, inside your project. Keep that folder. For documents with current
frozen results, the next project render reads those results instead of re-running the analysis. This
also lets another machine build those pages without running their R code. `_freeze/` is part of the
project, not a temporary folder to delete.
:::

### Discuss together: `cache` versus `freeze`

**Goal.** You have now met two things that both avoid re-running code. Explain how they differ and
when to use each one.

`cache: true` came up on the slides earlier today. You have just used `freeze: auto`. Both skip
work. They are not the same mechanism, and only one of them solves the problem you just solved.

::: {.callout-note appearance="simple" icon=false}
## {{< fa circle-check >}} You should be able to explain

| | `cache` | `freeze` |
|---|---|---|
| Scope | A cell within a document | A document within a project |
| Results | Local cache directory | `_freeze/` |
| Commit the results? | Normally no | Yes |
:::

::: {.callout-tip collapse="true"}
## Hint
Where does each one write its results, and would that location normally be committed? Those answers
explain their different purposes.

**`cache`** is knitr's chunk cache. It works *inside* one document, keyed on the chunk, and it is
about not re-running an expensive chunk while you iterate on that document. Its cache directory is
local scratch, and it is not something you ship.

**`freeze`** is a *project* policy. It stores each document's rendered results under `_freeze/`, which
you **do** commit, so a later build (including one on a machine with no R) reuses them.

So: `cache` for a slow chunk you are iterating on. `freeze` for making the project buildable without
re-running the science. `freeze: true` goes further than `auto`: a project build reuses a page's
stored results even after you edit it, so you decide when to refresh by rendering that page on its
own.
:::

### Put your Day-1 report on the site

*This is a core task.*

**Goal.** Take the report you wrote yesterday and make it a page of this website.

On Day 1, you built one document. Now add it to the project.

::: {.callout-note appearance="simple" icon=false}
## {{< fa circle-check >}} You should see
Your Day-1 report opens as a page of the site, reachable from the navigation, styled like the rest of
it. Its content is unchanged: the front matter is the only part you edited.
:::

::: {.callout-tip collapse="true"}
## Hint
Your Day-1 document may still declare its own `format:`. Inside a project, where does `format:` come
from now, and what happens if a page insists on its own?

Copy the `.qmd` into `day2-projects/`, together with `references.bib` and `apa.csl` from
`day1-intro/` (the front matter points at them by relative path), add an entry for it in the
`navbar:` list, and re-render.

A page in a project inherits `format:` from `_quarto.yml`. If your file still carries the
`format: typst` from the Day-1 exercise, it keeps building a PDF instead of a page. Drop that key and
it builds as HTML like its neighbours. (The shipped `penguins-report.qmd` declares no `format:` at
all, so there is nothing to drop.)

Didn't finish Day 1, or do not have the file? Use `solutions/day1/penguins-report.qmd` from
your exercises download instead.
:::

::: {.callout-note collapse="true"}
## Solution
Copy your `my-report.qmd` into `day2-projects/`, together with `references.bib` and `apa.csl` from
`day1-intro/` (the report cites through them). Drop any `format:` block from its front matter,
and add it to the navbar:

```{.yaml filename="day2-projects/_quarto.yml"}
website:
  navbar:
    left:
      - href: index.qmd
        text: Home
      - href: analysis.qmd
        text: Analysis
      - href: my-report.qmd
        text: My report
```

`solutions/day2/my-report.qmd` shows the finished page.
:::

### Guided experiment: `freeze: true`

*This step completes the core tasks. It is guided: run the commands as written.*

**Goal.** See what `freeze: true` changes compared to `auto`: the build reuses a page's stored
results even after you edit that page.

::: {.callout-note appearance="simple" icon=false}
## {{< fa circle-check >}} You should see
Check the label and timestamp after each render:

- **Project render:** both remain unchanged because the build reuses `_freeze/`.
- **Single-file render:** both update because the page executes.
- **Final project render:** the updated values remain because the build reuses the refreshed result.
:::

First, change `freeze` from `auto` to `true` in `_quarto.yml`:

```{.yaml filename="day2-projects/_quarto.yml"}
execute:
  freeze: true
```

Then, in `analysis.qmd`, make a visible edit to the plot code: change the y-axis label to
`y = "Body mass in grams"`.

Now run the three renders from inside `day2-projects/`, and reload `_site/analysis.html` in your
browser after each one:

```bash
quarto render                # whole project: your edit does not appear
quarto render analysis.qmd   # one file: the page refreshes
quarto render                # whole project: the refreshed page stays
```

With `freeze: true`, the workflow is: edit a page, render that file, then build the project without
re-running it. Switch back to `freeze: auto` before you continue. It is the setting to keep here.
Use `true` when project builds must reuse stored results until you deliberately refresh a page.

### Optional group demo: inspect a dashboard

If time permits, we will inspect this worked example together. You do not need to build a dashboard
during the workshop. We will open the [finished dashboard](dashboard.qmd){target="_blank"}, then
compare it with `solutions/day2/dashboard.qmd` to see how the same analysis becomes a dashboard.

Look for four ideas:

- `format: dashboard` changes the output.
- A `##` heading starts a row and a `###` heading starts a column.
- `#| content: valuebox` creates a single-number panel from a cell whose result is one number (for
  example `nrow(penguins)`).
- `{.tabset}` turns the cells in a column into tabs.

The overall shape is:

```markdown
---
title: "Penguins dashboard"
format: dashboard
---

## Row {height="20%"}

(two cells, each with `#| content: valuebox`)

## Row {height="80%"}

### Column

(the boxplot cell, with a `#| title:`)

### Column {.tabset}

(the scatter cell and the table cell: each becomes a tab)
```

After the group demo, you can optionally copy `analysis.qmd` to `dashboard.qmd` and reproduce one
part of the layout.

### Put it online *(optional, and a good place to stop)*

**Goal.** Take the `_site/` folder you have built and put it on a URL you can open on your phone.

::: {.column-margin}
{{< fa book-open >}} **Quarto docs**

[Publish to Posit Connect Cloud](https://quarto.org/docs/publishing/posit-connect-cloud.html)
:::

`_site/` **is** the core deliverable. Any static host will serve it: an internal server, your
department's pages, or a hosting service. Publishing here is optional. If you made a free Posit
Connect Cloud account before the workshop, one command puts the site online.

::: {.callout-note appearance="simple" icon=false}
## {{< fa circle-check >}} You should see
A browser opens once to sign you in. Then the contents of `_site/` upload, and you get back a URL
that serves your site.
:::

::: {.callout-note collapse="true"}
## The command
From inside your project folder:

```bash
quarto publish posit-connect-cloud
```

No repository is needed. Connect Cloud takes the locally rendered files as they are.

No account, or the sign-in is blocked on your machine? You still have `_site/`, which was the actual
deliverable. Nothing downstream depends on this step.
:::

::: {.callout-note}
## For later: package versions
`renv::snapshot()` writes an `renv.lock` that pins your package versions. That is the second half of
reproducibility: `freeze` preserves the *results*, `renv.lock` preserves the *environment* that
produced them.
:::

## Troubleshooting

::: {.callout-tip collapse="true" icon=false}
## {{< fa wrench >}} If rendering fails or the output looks wrong
- **Rendered, but can't find the HTML?** Before you add `_quarto.yml`, a single-page render writes
  the HTML **next to its source** in `day2-projects/`. Once `_quarto.yml` is present, a project render
  collects all pages under `day2-projects/_site/`. Look in those two places. The earlier
  `analysis.html` and `analysis_files/` next to the sources are leftovers at that point: delete them
  if you like.
- **YAML is indentation-sensitive.** If rendering fails *before* any code runs, check `_quarto.yml`
  for stray spaces, tabs, or unmatched quotes.
- **"No project" / pages don't link?** The `_quarto.yml` must be **inside `day2-projects/`**, and
  `href:` paths are relative to the source folder, **not** `_site/`.
- **Brand not applied?** Two causes. The file must be in `day2-projects/` (the project root), and
  `brand` must come **last** in the `theme:` list: `theme: [cosmo, brand]` gives you the palette,
  while `theme: cosmo` on its own keeps cosmo's colors and only the fonts change. Re-render after
  either fix. For
  **plots**, the palette needs `library(brand.yml)` + `theme_brand_ggplot2()` (the file alone doesn't
  reach ggplot).
- **Render fails with "references.bib not found"?** Your Day-1 report points at `references.bib` and
  `apa.csl` by relative path. Copy both files from `day1-intro/` into `day2-projects/`, next to the
  report, and render again.
- **A cross-reference shows `?@…`?** The cell label must start `fig-`/`tbl-`/`eq-` and be unique, and
  remember it only resolves **within one page** on a website.
- **Freeze didn't skip the cell?** Under `freeze: auto` it skips only when the source is **unchanged**
  since the last build: any edit (prose or code) re-runs that page. Delete `_freeze/` to force a
  clean rebuild.
- **A page won't update?** That is `freeze: true` doing its job: a project render reuses the stored
  results even after an edit. Render that file directly, or switch back to `freeze: auto`.
- **Missing package?** `install.packages("<name>")` (or `renv::restore()`) and render again.
:::

## Session {.appendix .unnumbered}

<details>
<summary>Session info</summary>

```{r}
#| label: session
sessionInfo()
```

</details>
