---
title: "Quarto projects — Lab"
subtitle: "RaukR 2026 · Day 2"
author: ""
date: ""
format:
  html:
    toc-location: left   # free the right margin for the "Quarto docs" pointers
---

```{r}
#| label: setup
#| message: false
# Renders the target figure shown further down this page — you'll do the actual
# work inside your day2-projects/ folder, not here. This cell is not step zero.
library(dplyr)
library(ggplot2)
data(penguins)
penguins <- penguins |> filter(!is.na(bill_len), !is.na(bill_dep), !is.na(body_mass))
```

::: {.callout-note}
## Scope
Turn a set of `.qmd` into a Quarto **project**: a navigable, **branded website**, then make its
builds reproducible with **freeze** and render it to a publishable folder. Two challenges, one
project: each is its own hands-on part, with a break between them. Putting the site online is an
**optional last step** (Posit Connect Cloud, free account, no repository), so the folder is what
everyone leaves with. Everything runs on base-R `datasets::penguins` (R ≥ 4.5), so there's no data to download.

**Before you start:** work through the **[Setup page](../../setup.qmd)** if you haven't. Its
`00-check-setup.R` confirms R, Quarto, and the packages this lab uses. The optional branded-plot
stretch also uses `brand.yml`.
:::

**Starting point: a shipped set of pages.** You don't start from a blank folder. Open your
**`day2-projects/`** folder from the exercises download (double-click `day2-projects.Rproj`, see the
[Setup page](../../setup.qmd#get-the-materials)). It holds four content pages that already render on
their own but are **not yet a project**:

- `index.qmd` — a home page
- `analysis.qmd` — a penguins analysis with a figure and a table
- `reports/methods.qmd` and `reports/data-notes.qmd` — two short supporting pages, in their own
  folder because you will treat that folder differently later on

Render one page to confirm it works: `quarto render analysis.qmd` gives a normal HTML page
(`analysis.html` right next to its source), since the folder isn't a project *yet*. Your job across the
two challenges is to make it a *site*, then ship it. Everything you render lands inside
`day2-projects/`. There's no project folder above it, so nothing redirects your output. A finished
reference is in **`solutions/day2/`** if you get stuck or fall behind.

## Website Challenge

> **Goal:** turn the `day2-projects/` pages into a navigable, **branded website** that renders locally to a
> `_site/` folder.

Five steps, each one a task rather than a block of YAML to copy. Every step states what you should
see *before* you start, so you can tell for yourself when you are done. If you get stuck, open the
hints in order: the first one is a question to ask yourself, the second names the mechanism, and the
last one is the answer.

::: {.column-margin}
{{< fa book-open >}} **Quarto docs**

[Websites](https://quarto.org/docs/websites/) · [Projects](https://quarto.org/docs/projects/quarto-projects.html) · [Brand](https://quarto.org/docs/authoring/brand.html)
:::

**The `solutions/day2/` folder is a reference, not a checklist.** The complete, known-good project
lives there in your exercises download. Open it to compare at any point, or copy it wholesale if you
fall behind.

### Make it a project

**Goal.** Make Quarto treat `day2-projects/` as one project, so that a single command builds every
page in it into an output folder.

Here is the stanza that declares a project. Where it goes, and how you then build the folder, is the
work:

```yaml
project:
  type: website
  output-dir: _site
```

::: {.callout-note appearance="simple" icon=false}
## {{< fa circle-check >}} You should see
One command, run with **no file argument**, builds **both** pages at once, and the HTML lands in a
new `day2-projects/_site/` folder instead of next to each `.qmd`. Before this step, rendering
`analysis.qmd` put `analysis.html` right beside it.
:::

::: {.callout-tip collapse="true"}
## Stuck? (1) — a question to ask yourself
You have folders above `day2-projects/` and files inside it. Which one does Quarto look at to decide
where a project starts, and in which direction does it search?
:::

::: {.callout-tip collapse="true"}
## Stuck? (2) — the mechanism
Quarto walks **up** from the file being rendered and stops at the **nearest** `_quarto.yml`. That
file marks the project root. Nothing sits above `day2-projects/` in your download, so a `_quarto.yml`
placed inside it makes that folder the project and nothing else can claim your pages.

A `quarto render` with **no file argument** builds the whole project. `quarto render index.qmd`
still builds just the one page.
:::

::: {.callout-note collapse="true"}
## Solution
Create `day2-projects/_quarto.yml`:

```{.yaml filename="day2-projects/_quarto.yml"}
project:
  type: website
  output-dir: _site
```

Then, from inside `day2-projects/`:

```bash
quarto render
```
:::

### Build the navigation

**Goal.** Give the site a title and a top navigation bar, so a reader can move between the two pages
without editing the URL.

Your pages are `index.qmd` (Home) and `analysis.qmd` (Analysis). Give the site the title
**Penguin Lab**, and use the ready-made **cosmo** theme so it does not look like raw HTML.

::: {.callout-note appearance="simple" icon=false}
## {{< fa circle-check >}} You should see
Opening `_site/index.html` shows the home page with a bar across the top carrying the site title and
two entries, **Home** and **Analysis**. Clicking Analysis moves you to the other page, which carries
this figure:
:::

```{r}
#| label: fig-target
#| echo: false
#| fig-width: 6
#| fig-height: 3.2
#| fig-cap: "The analysis page's figure: body mass by species."
#| fig-alt: >-
#|   Boxplot of penguin body mass in grams for Adelie, Chinstrap and Gentoo species; Gentoo is
#|   clearly the heaviest, Adelie and Chinstrap overlap.
ggplot(penguins, aes(species, body_mass, fill = species)) +
  geom_boxplot(show.legend = FALSE) +
  labs(x = NULL, y = "Body mass (g)")
```

::: {.callout-tip collapse="true"}
## Stuck? (1) — a question to ask yourself
The `project:` key configures the *build*. A navbar is not about building, it is about the site
itself. What would you expect that second top-level key to be called, and what does each navbar entry
need to know?
:::

::: {.callout-tip collapse="true"}
## Stuck? (2) — the mechanism
A `website:` key sits next to `project:` at the top level. It takes a `title:`, and a `navbar:` with a
`left:` (or `right:`) list. Each entry is a pair: `href:` pointing at the **source** `.qmd` (not the
built `.html`), and `text:` for the label.

The theme is a **format** option, so it goes under `format: html:`.
:::

::: {.callout-note collapse="true"}
## Solution
```{.yaml filename="day2-projects/_quarto.yml"}
project:
  type: website
  output-dir: _site

website:
  title: "Penguin Lab"
  navbar:
    left:
      - href: index.qmd
        text: Home
      - href: analysis.qmd
        text: Analysis

format:
  html:
    theme: cosmo
```
:::

::: {.callout-tip collapse="true"}
## Try the other menu: a sidebar built for you
Swap the `navbar:` block for this, re-render, and compare:

```yaml
  sidebar:
    contents: auto
```

`contents: auto` builds the menu from the files it finds, so you never list pages by hand. It is the
right choice for a site that grows. A navbar is the right choice when you want to control exactly
what appears and in what order. Put the navbar back before you carry on, or keep both.
:::

::: {.callout-note}
## Linking pages, and what does not carry across
Two things here are worth a moment, because they are the Day-1 rules meeting a multi-page site.

`index.qmd` already links to the analysis with a plain Markdown link, `[analysis](analysis.qmd)`.
That is deliberate. **Across** pages there is no auto-numbering (that is a *book* feature, not a
website one), so a plain link is how you connect website pages.

**Within** one page, yesterday's rule still holds. Drop `@tbl-means` into a sentence on
`analysis.qmd` and it resolves to *Table 1*, because the reference and its table share one page.
:::

### Brand it

**Goal.** Give the site one shared palette and typeface from a `_brand.yml`, the same file you met on
Day 1.

Here is the brand body. Getting it to actually take effect is the work:

```{.yaml filename="day2-projects/_brand.yml"}
color:
  palette:
    teal: "#4C979F"
  primary: teal
typography:
  fonts:
    - family: Albert Sans
      source: google
  base:
    family: Albert Sans
```

::: {.callout-note appearance="simple" icon=false}
## {{< fa circle-check >}} You should see
**First, a surprise, and it is meant to happen.** Add the file, re-render, and the **type** changes
(the site is set in Albert Sans) while the **colors do not**: the navbar and links stay cosmo blue.
Nothing is broken and you have not mistyped anything. Working out why is this step.

**When you have solved it**, the navbar and the links are teal. Headings keep the body color the
whole time, so do not use them to check: `primary` colors the navbar and links, not headings. The
plot keeps its default species colors too, because branding an R plot is a separate
`brand.yml`-package step, out of scope today.
:::

::: {.callout-tip collapse="true"}
## Stuck? (1) — a question to ask yourself
`theme:` takes a **list**, even when you write a single value. So Quarto is already combining
something with your `cosmo`. If two layers both define a primary color, which one would you expect to
end up on the page?
:::

::: {.callout-tip collapse="true"}
## Stuck? (2) — the mechanism
`brand` is a theme **layer**, and it is the layer that reads your `_brand.yml`. A bare
`theme: cosmo` expands to `[brand, cosmo]`, and **later layers win**, so cosmo's palette lands on top
of the brand's and you get cosmo's blue.

The fonts come through whichever way round you write it, which is exactly what makes this look like
it worked.
:::

::: {.callout-note collapse="true"}
## Solution
Put `brand` **last** in the list:

```{.yaml filename="day2-projects/_quarto.yml"}
format:
  html:
    theme: [cosmo, brand]
```

Dropping `theme:` altogether works too: with no named theme to fight, the brand is all there is.
:::

### Scope options to one folder

**Goal.** Fold the code on the two pages in `reports/`, and only there. The home and analysis pages
keep showing their code as before.

You have not looked at `day2-projects/reports/` yet. It holds two short pages, `methods.qmd` and
`data-notes.qmd`, each with a small code cell.

::: {.callout-note appearance="simple" icon=false}
## {{< fa circle-check >}} You should see
On the two `reports/` pages, each code cell is collapsed behind a **Show the code** toggle that
expands when clicked. On `index.qmd` and `analysis.qmd`, nothing changes. You did not edit any of the
four `.qmd` files to get this.
:::

::: {.callout-tip collapse="true"}
## Stuck? (1) — a question to ask yourself
Project-wide options live in `_quarto.yml` at the project root. You want the same kind of option, but
reaching one folder only. Where would you put a file so that Quarto applies it to that folder and
nowhere else?
:::

::: {.callout-tip collapse="true"}
## Stuck? (2) — the mechanism
A `_metadata.yml` inside a folder applies to every document in that folder (and below it). Quarto
merges it on top of the project defaults, so you write only what differs.

It holds **document** options directly, without the `format:` wrapper you would need in
`_quarto.yml`. The option that folds code is `code-fold`, and `code-summary` sets the toggle label.
:::

::: {.callout-note collapse="true"}
## Solution
Create `day2-projects/reports/_metadata.yml`:

```{.yaml filename="day2-projects/reports/_metadata.yml"}
code-fold: true
code-summary: "Show the code"
```

Re-render the project. Nothing else changes.
:::

### Index the reports *(stretch)*

**Goal.** Put a table of the `reports/` pages on the home page, built by Quarto rather than typed by
you, so that adding a page there shows up without editing `index.qmd`.

Docs: [Website listings](https://quarto.org/docs/websites/website-listings.html).

::: {.callout-note appearance="simple" icon=false}
## {{< fa circle-check >}} You should see
A table on the home page with one row per page in `reports/`, showing each title and its description
(both pages already carry a `description:` in their front matter). Add a third page to `reports/`,
re-render, and it appears in the table on its own.
:::

::: {.callout-tip collapse="true"}
## Stuck? (1) — a question to ask yourself
This is a property of one page, not of the whole site. So where does its configuration go, and how
does the page say *which* documents to index?
:::

::: {.callout-tip collapse="true"}
## Stuck? (2) — the mechanism
A `listing:` key in the **front matter of `index.qmd`** points at a folder with `contents:`. Give it
an `id:`, then place the result where you want it on the page with an empty div carrying that id:

```markdown
::: {#reports}
:::
```

Without the `id:` and the div, the listing renders at the bottom of the page.
:::

::: {.callout-note collapse="true"}
## Solution
In the front matter of `day2-projects/index.qmd`:

```{.yaml filename="day2-projects/index.qmd"}
---
title: "Penguin Lab"
listing:
  id: reports
  contents: reports
  type: table
  fields: [title, description]
  sort: title
---
```

Then, in the body, where you want the table:

```markdown
## Reports

::: {#reports}
:::
```
:::

## Ship it Challenge

> **Goal:** make the project's builds **reproducible**, bring your own work onto the site, and render
> it all to a publishable `_site/` folder.

::: {.callout-note}
## Starting point
Keep going with your Website Challenge `day2-projects/`. If you didn't finish it, open the ready-made
**`solutions/day2/`** project (a complete Part-1 site) and do the steps below in there.
:::

Same shape as the first part: a task, then what you should see, then hints in order if you want them.

::: {.column-margin}
{{< fa book-open >}} **Quarto docs**

[Code execution](https://quarto.org/docs/projects/code-execution.html) · [Dashboards](https://quarto.org/docs/dashboards/) · [Publishing](https://quarto.org/docs/publishing/)
:::

### Make the build skip unchanged pages

**Goal.** Right now every `quarto render` re-runs all the R on every page, even the pages you have not
touched. Make Quarto reuse the previous results for a page whose source has not changed. Then
**prove** that it worked.

The proof is half the task, and it is the more interesting half. An absence in the log is weak
evidence. Find something you can point at.

::: {.callout-note appearance="simple" icon=false}
## {{< fa circle-check >}} You should see
Render twice with no edits in between. On the second render, the R on your pages does not run again,
and you can **show** it: something on the rendered page that would certainly be different if the code
had run, and is not. Then edit the code on one page only, render again, and see that page (and only
that page) come back to life.
:::

::: {.callout-tip collapse="true"}
## Stuck? (1) — a question to ask yourself
Two questions, one for each half. For the setting: this is about *executing* code across the whole
project, so which top-level key would carry it? For the proof: what could a cell print that is
**guaranteed** to differ between two runs a minute apart?
:::

::: {.callout-tip collapse="true"}
## Stuck? (2) — the mechanism
An `execute:` key at the top level of `_quarto.yml` carries execution options for the project. The one
you want is `freeze`, and `auto` means "re-execute only when this document's own source changes".

For the proof: print the current time. If the value on the page is identical across two renders, the
code that produced it did not run.
:::

::: {.callout-note collapse="true"}
## Solution
In `day2-projects/_quarto.yml`:

```{.yaml filename="day2-projects/_quarto.yml"}
execute:
  freeze: auto
```

For the visible tell, add a cell to `analysis.qmd`:

```r
cat(format(Sys.time()))
```

Then, from inside `day2-projects/`:

```bash
quarto render     # runs the code, writes _freeze/ and _site/
quarto render     # no edits: cells are skipped, and the printed time is unchanged
```

Now change the plot code on `analysis.qmd` and render once more. Only that page re-runs, and its
timestamp finally moves.
:::

::: {.callout-note}
## Why this matters beyond the wait
The frozen results live in `_freeze/`. Commit that folder and a machine with **no R at all** can
rebuild the site from it, which is how an automated build serves a site whose figures came out of R.
That is the reason to keep `_freeze/` in version control rather than treat it as scratch.
:::

### `cache` versus `freeze`

**Goal.** You have now met two things that both avoid re-running code. Work out how they differ, and
when you would reach for each.

You met `cache: true` on Day 1. You have just used `freeze: auto`. Both skip work. They are not the
same mechanism, and only one of them solves the problem you just solved.

::: {.callout-note appearance="simple" icon=false}
## {{< fa circle-check >}} You should see
You can answer three questions without looking them up again: which one is set per document and which
is a project-level build policy; which one stores results that travel with the repository for another
machine to reuse; and which one would still re-run everything when a colleague clones the project
fresh.
:::

::: {.callout-tip collapse="true"}
## Stuck? (1) — a question to ask yourself
Where does each one write its results, and would that location normally be committed? Follow that and
the rest falls out.
:::

::: {.callout-tip collapse="true"}
## Stuck? (2) — the answer
**`cache`** is knitr's chunk cache. It works *inside* one document, keyed on the chunk, and it is
about not re-running an expensive chunk while you iterate on that document. Its cache directory is
local scratch, and it is not something you ship.

**`freeze`** is a *project* policy. It stores each document's rendered results under `_freeze/`, which
you **do** commit, so a later build (including one on a machine with no R) reuses them.

So: `cache` for a slow chunk you are iterating on. `freeze` for making the project buildable without
re-running the science. `freeze: true` goes further than `auto` and never re-executes on a project
build, which is the setting an automated build uses.
:::

### Put your Day-1 report on the site

**Goal.** Take the report you wrote yesterday and make it a page of this website.

This is the whole point of the two days. Day 1 you built one document. Today that document becomes
part of something bigger, and it should take almost nothing to do.

::: {.callout-note appearance="simple" icon=false}
## {{< fa circle-check >}} You should see
Your Day-1 report opens as a page of the site, reachable from the navigation, styled like the rest of
it. Its content is unchanged. The only thing you had to reconsider was its front matter.
:::

::: {.callout-tip collapse="true"}
## Stuck? (1) — a question to ask yourself
Your Day-1 document set its own `format:`. Inside a project, where does `format:` come from now, and
what happens if the page keeps insisting on its own?
:::

::: {.callout-tip collapse="true"}
## Stuck? (2) — the mechanism
Copy the `.qmd` into `day2-projects/`, add an entry for it in the `navbar:` list, and re-render.

A page in a project inherits `format:` from `_quarto.yml`, so the `format: pdf` from Day 1 is no
longer what you want here. Drop it and the page builds as HTML like its neighbours.

Didn't finish Day 1, or don't have the file to hand? Use `solutions/day1/penguins-report.qmd` from
your exercises download instead.
:::

::: {.callout-note collapse="true"}
## Solution
Copy your `my-report.qmd` into `day2-projects/`, remove the `format:` block from its front matter,
and add it to the navbar:

```{.yaml filename="day2-projects/_quarto.yml"}
website:
  navbar:
    left:
      - href: index.qmd
        text: Home
      - href: analysis.qmd
        text: Analysis
      - href: my-report.qmd
        text: My report
```

`solutions/day2/my-report.qmd` shows the finished page.
:::

### Ship a dashboard

**Goal.** Turn the analysis into a **dashboard**: the same numbers, arranged as a layout of panels
rather than a page you scroll.

Start from a copy of `analysis.qmd` (call it `dashboard.qmd`) so you are rearranging content you
already have rather than writing new analysis. Aim for: a row of two summary figures across the top,
and a wider row below holding the boxplot next to a tabbed panel with the scatter plot and the table.

::: {.callout-note appearance="simple" icon=false}
## {{< fa circle-check >}} You should see
A page that fills the browser window instead of scrolling. Across the top, two small panels each
showing a single number (how many species, how many penguins). Below, the boxplot in its own titled
panel, and beside it a panel with **tabs** you can click between to swap the scatter plot for the
table. Resizing the window rearranges it rather than clipping it.
:::

::: {.callout-tip collapse="true"}
## Stuck? (1) — a question to ask yourself
Nothing about your R changes here, so what changes? And once you have said "this is a dashboard", how
would a Markdown document express *rows* and *columns* using structure it already has?
:::

::: {.callout-tip collapse="true"}
## Stuck? (2) — the mechanism
`format: dashboard` in the front matter. That is the only change to the document's setup.

The layout comes from **headings**: a `##` starts a row, a `###` starts a column inside it. The
heading text is only a label. Each code cell becomes a panel, and `#| title:` sets its header.

For the single-number panels, `#| content: valuebox` turns a cell into one, and the cell returns a
`list(color =, icon =, value =)`. For the tabs, put `{.tabset}` on the column heading, and each cell
in that column becomes a tab. Row heights are set with `{height="20%"}` on the row heading.
:::

::: {.callout-note collapse="true"}
## Solution
`solutions/day2/dashboard.qmd` is the finished dashboard, with comments marking each part of the
layout. The shape is:

```markdown
---
title: "Penguins dashboard"
format: dashboard
---

## Row {height="20%"}

(two cells, each with `#| content: valuebox`)

## Row {height="80%"}

### Column

(the boxplot cell, with a `#| title:`)

### Column {.tabset}

(the scatter cell and the table cell: each becomes a tab)
```
:::

### Put it online *(optional, and a good place to stop)*

**Goal.** Take the `_site/` folder you have built and put it on a URL you can open on your phone.

`_site/` **is** the deliverable. Any static host will serve it: an internal server, your department's
pages, a hosting service. If you made a free Posit account before the workshop, one command does it
from here.

::: {.callout-note appearance="simple" icon=false}
## {{< fa circle-check >}} You should see
A browser opens once to sign you in. Then the contents of `_site/` upload, and you get back a URL
that serves your site.
:::

::: {.callout-note collapse="true"}
## The command
From inside your project folder:

```bash
quarto publish posit-connect-cloud
```

No repository is needed. Connect Cloud takes the locally rendered files as they are.

No account, or the sign-in is blocked on your machine? You still have `_site/`, which was the actual
deliverable. Nothing downstream depends on this step.
:::

::: {.callout-note}
## One more leg, for later
`renv::snapshot()` writes an `renv.lock` that pins your package versions. That is the second half of
reproducibility: `freeze` preserves the *results*, `renv.lock` preserves the *environment* that
produced them.
:::

## Troubleshooting

::: {.callout-tip collapse="true" icon=false}
## {{< fa wrench >}} If a render misbehaves
- **Rendered, but can't find the HTML?** Before you add `_quarto.yml`, a single-page render writes
  the HTML **next to its source** in `day2-projects/`. Once `_quarto.yml` is present, a project render
  collects both pages under `day2-projects/_site/`. Look in those two places.
- **YAML is indentation-sensitive.** If rendering fails *before* any code runs, check `_quarto.yml`
  for stray spaces, tabs, or unmatched quotes.
- **"No project" / pages don't link?** The `_quarto.yml` must be **inside `day2-projects/`**, and
  `href:` paths are relative to the source folder, **not** `_site/`.
- **Brand not applied?** Two causes. The file must be in `day2-projects/` (the project root), and
  `brand` must come **last** in the `theme:` list: `theme: [cosmo, brand]` gives you the palette,
  while `theme: cosmo` on its own keeps cosmo's colors and only the fonts change. Re-render after
  either fix. For
  **plots**, the palette needs `library(brand.yml)` + `theme_brand_ggplot2()` (the file alone doesn't
  reach ggplot).
- **A cross-reference shows `?@…`?** The cell label must start `fig-`/`tbl-`/`eq-` and be unique, and
  remember it only resolves **within one page** on a website.
- **Freeze didn't skip the cell?** It skips only when the source is **unchanged** since the last build:
  any edit (prose or code) re-runs that page. Delete `_freeze/` to force a clean rebuild.
- **Missing package?** `install.packages("<name>")` (or `renv::restore()`) and render again.
:::

## Session {.appendix .unnumbered}

<details>
<summary>Session info</summary>

```{r}
#| label: session
sessionInfo()
```

</details>
