---
title: "Quarto projects — Lab"
subtitle: "RaukR 2026 · Day 2"
author: ""
date: ""
---

```{r}
#| label: setup
#| message: false
# Renders the target figure shown further down this page — you'll do the actual
# work inside your day2-projects/ folder, not here. This cell is not step zero.
library(dplyr)
library(ggplot2)
data(penguins)
penguins <- penguins |> filter(!is.na(bill_len), !is.na(bill_dep), !is.na(body_mass))
```

::: {.callout-note}
## Scope
Turn a set of `.qmd` into a Quarto **project**: a navigable, **branded website**, then make its
builds reproducible with **freeze** and render it to a publishable folder. Two challenges, one
project — each is its own hands-on part, with a break between them. Publishing to
GitHub Pages / CI is a **watch-me demo**, not a per-participant step — publishing needs an account and
auth setup, so we run it once up front. Everything runs on base-R `datasets::penguins` (R ≥ 4.5) — no data to download.

**Packages:** `dplyr`, `ggplot2`, plus the `knitr`/`rmarkdown` engine every `.qmd` with R code needs.
If you've done the **[Setup page](../../setup.qmd)** (with its `00-check-setup.R` check), you're
ready; otherwise install them once from R: `install.packages(c("dplyr", "ggplot2"))`. Optional branded
plots use the **`brand.yml`** package (`install.packages("brand.yml")`).
:::

**Starting point — a shipped set of pages.** You don't start from a blank folder. Open your
**`day2-projects/`** folder from the exercises download (double-click `day2-projects.Rproj` — see the
[Setup page](../../setup.qmd#get-the-materials)). It holds two content pages that already render on
their own but are **not yet a project**:

- `index.qmd` — a home page;
- `analysis.qmd` — a penguins analysis with a figure and a table.

Render one page to confirm it works: `quarto render analysis.qmd` gives a normal HTML page —
`analysis.html` right next to its source, since the folder isn't a project *yet*. Your job across the
two challenges is to make it a *site*, then ship it. Everything you render lands inside
`day2-projects/`; there's no project folder above it, so nothing redirects your output. A finished
reference is in **`solutions/day2/`** if you get stuck or fall behind — nobody is stranded by the break.

## Website Challenge

> **Goal:** turn the `day2-projects/` pages into a navigable, **branded website** that renders locally to a
> `_site/` folder.

::: {.callout-note icon=false}
## {{< fa clipboard-list >}} Tasks
1. In `day2-projects/`, create a **`_quarto.yml`** that declares the project and a website with navigation:
   ```yaml
   project:
     type: website
     output-dir: _site
   website:
     title: "Penguin Lab"
     navbar:
       left:
         - href: index.qmd
           text: Home
         - href: analysis.qmd
           text: Analysis
   format:
     html:
       theme: cosmo
   ```
2. **Render the whole folder** — from inside `day2-projects/`, run `quarto render` (or `quarto preview`).
   Because a `_quarto.yml` is now present, Quarto treats the folder as one project and builds **both**
   pages into `_site/`.
3. **Brand it.** Add a `day2-projects/_brand.yml` so the site shares one palette:
   ```yaml
   color:
     palette:
       teal: "#4C979F"
     primary: teal
   typography:
     fonts:
       - family: Albert Sans
         source: google
     base:
       family: Albert Sans
   ```
   Re-render — the navbar and headings pick up the palette automatically.
4. *(stretch)* Two things make the **within-page versus across-page** rule concrete. First, on
   `analysis.qmd`, drop `@tbl-means` into a sentence — it resolves to *Table 1* because the reference
   and its table share **one page** (exactly as yesterday). Second, notice `index.qmd` **already**
   links to the analysis with a plain Markdown link — `[analysis](analysis.qmd)`: that's deliberate,
   because **across** pages there is no auto-numbering (a *book* feature, not a website one), so a
   plain link is how you connect website pages.
:::

::: {.callout-note appearance="simple" icon=false}
## {{< fa circle-check >}} You should see
A `day2-projects/_site/` folder with **two linked pages** and a **top navbar** (Home · Analysis); after the
brand step, the navbar and headings turn **teal** with Albert Sans (the plot keeps its default species
colours — branding an R plot is a separate `brand.yml`-package step, out of scope today). If you did the stretch,
the analysis page also shows a live **Table 1** reference (from `@tbl-means`) and the home page carries
a plain link into the analysis. Opening `_site/index.html` shows the home page; the navbar moves you to
the analysis page, whose figure is:
:::

```{r}
#| label: fig-target
#| echo: false
#| fig-width: 6
#| fig-height: 3.2
#| fig-cap: "The analysis page's figure — body mass by species."
#| fig-alt: >-
#|   Boxplot of penguin body mass in grams for Adelie, Chinstrap and Gentoo species; Gentoo is
#|   clearly the heaviest, Adelie and Chinstrap overlap.
ggplot(penguins, aes(species, body_mass, fill = species)) +
  geom_boxplot(show.legend = FALSE) +
  labs(x = NULL, y = "Body mass (g)")
```

::: {.callout-tip collapse="true"}
## Hint
- The `_quarto.yml` must sit **inside `day2-projects/`** — Quarto uses the *nearest* `_quarto.yml` as
  the project root, so this makes `day2-projects/` its own project. There's no `_quarto.yml` in any
  folder above it, so nothing else can claim your pages.
- `navbar:` items are `href` + `text`; `sidebar: { contents: auto }` builds a menu from the folder's
  files instead.
- `_brand.yml` is **auto-discovered** at the project root — no header line needed; just re-render.
- `quarto render` with no file argument builds the **whole project**; `quarto render index.qmd`
  builds one page.
:::

**The `solutions/day2/` folder is a reference, not a checklist.** The complete, known-good project lives
there in your exercises download: open it to compare your `_quarto.yml` / `_brand.yml`, or copy it
wholesale if you fell behind.

## Ship it Challenge

> **Goal:** make the project's builds **reproducible** with `freeze`, then render it to a publishable
> `_site/` folder. (Publishing to Pages is the watch-me demo — you produce the folder.)

::: {.callout-note}
## Starting point
Keep going with your Website Challenge `day2-projects/`. If you didn't finish it, open the ready-made
**`solutions/day2/`** project — a complete Part-1 site — and do the steps below in there. Nobody is stranded.
:::

::: {.callout-note icon=false}
## {{< fa clipboard-list >}} Tasks
1. Add **freeze** to your project's `_quarto.yml` so unchanged pages aren't re-executed on every build:
   ```yaml
   execute:
     freeze: auto      # re-execute a document only when its own source changes
   ```
2. **See the skip.** Run `quarto render` once (the code runs; `_freeze/` is written). Run
   `quarto render` **again without editing** — the R cell is **skipped**, its result reused from
   `_freeze/` (the log shows no compute). For a *visible* tell, drop a cell with
   `cat(format(Sys.time()))` on `analysis.qmd`: on the frozen second render the printed timestamp does
   **not** change — a pass/fail you can see, not just an absence in the log. Now change the plot's
   **code** on `analysis.qmd` and re-render — only *that* page re-runs (and the timestamp finally ticks).
3. Find the shipped site under **`_site/`** — that folder *is* what you publish. Point any static host
   at it (GitHub Pages, Netlify, an internal server).
4. *(note, not a step)* `renv::snapshot()` writes an `renv.lock` that pins the packages — the second
   reproducibility leg, so a teammate (or CI) rebuilds the same environment.
:::

::: {.callout-note appearance="simple" icon=false}
## {{< fa circle-check >}} You should see
On the **second** render (no edits between them), the log shows the R cell **skipped** — reused from
`_freeze/`, and the `cat(format(Sys.time()))` line prints the **same** time it did on the first render.
That frozen timestamp (and the quiet log) is exactly what lets CI render the site with no R at all;
edit a page's code, and only that page re-runs — its timestamp finally advancing.
:::

::: {.callout-tip collapse="true"}
## Hint
- `freeze: auto` re-executes a document when its **source** changes (prose or code) and reuses
  `_freeze/` when it hasn't — so an *unchanged* page never re-runs. `freeze: true` goes further: never
  re-execute on a project build (the CI mode).
- Frozen results live in `_freeze/` — commit that folder so CI can render without R.
- `output-dir` (default `_site`) is the whole deliverable: publish the folder, don't rebuild on the host.
:::

```{r}
#| label: sol-ship
#| code-fold: true
#| code-summary: "Solution — the freeze block and the render flow"
#| eval: false
# add to your project's _quarto.yml:
execute:
  freeze: auto

# optional visible tell — put this in a cell on analysis.qmd:
#   cat(format(Sys.time()))       # frozen render reuses the old value, so the time stays put

# from inside your project folder (day2-projects/):
#   quarto render                 # 1st build: runs the code, writes _freeze/ and _site/
#   quarto render                 # 2nd build, no edits: cell skipped, timestamp unchanged
#   # ...change the plot's code on analysis.qmd, render again: only that page re-runs (time ticks)...
#
# publish = point a host at _site/  (watch-me: quarto publish gh-pages, auth pre-flighted)
```

## Troubleshooting

::: {.callout-tip collapse="true" icon=false}
## {{< fa wrench >}} If a render misbehaves
- **Rendered, but can't find the HTML?** Before you add `_quarto.yml`, a single-page render writes
  the HTML **next to its source** in `day2-projects/`; once `_quarto.yml` is present, a project render
  collects both pages under `day2-projects/_site/`. Look in those two places.
- **YAML is indentation-sensitive.** If rendering fails *before* any code runs, check `_quarto.yml`
  for stray spaces, tabs, or unmatched quotes.
- **"No project" / pages don't link?** The `_quarto.yml` must be **inside `day2-projects/`**, and
  `href:` paths are relative to the source folder, **not** `_site/`.
- **Brand not applied?** `_brand.yml` must be in `day2-projects/` (the project root); re-render after adding it. For
  **plots**, the palette needs `library(brand.yml)` + `theme_brand_ggplot2()` — the file alone doesn't
  reach ggplot.
- **A cross-reference shows `?@…`?** The cell label must start `fig-`/`tbl-`/`eq-` and be unique — and
  remember it only resolves **within one page** on a website.
- **Freeze didn't skip the cell?** It skips only when the source is **unchanged** since the last build
  — any edit (prose or code) re-runs that page. Delete `_freeze/` to force a clean rebuild.
- **Missing package?** `install.packages("<name>")` (or `renv::restore()`) and render again.
:::

## Session {.appendix .unnumbered}

<details>
<summary>Session info</summary>

```{r}
#| label: session
sessionInfo()
```

</details>
