---
title: "Penguin bill dimensions"
subtitle: "Part-2 starter — a finished Part-1 document, ready to cite"
author: "Your name"          # ← replace with your own before rendering
format: html
---

<!--
This is the KNOWN-GOOD Part-2 starter (running-order rule 2): a complete Part-1 penguins report —
a cross-referenced figure and table, a margin element, and a display equation — with NO citations
yet. Open it at the start of Part 2 (whether or not you finished your own Part-1 document) and add
the citation + Typst payoff on top of it. The fully cited version is `penguins-report.qmd`.
-->

```{r}
#| label: setup
#| message: false
library(dplyr)
library(ggplot2)
library(gt)
library(ggokabeito)   # color-blind-safe (Okabe-Ito) scale for the species colors

# Base-R penguins — zero install on R >= 4.5. Drop rows with missing bill measurements.
data(penguins)
penguins <- penguins |> filter(!is.na(bill_len), !is.na(bill_dep))
```

## The data

We measured **`{r} nrow(penguins)`** penguins of `{r} length(unique(penguins$species))` species
(`{r} knitr::combine_words(levels(penguins$species))`), collected at Palmer Station, Antarctica. The
species are not evenly represented:

```{r}
#| label: counts
#| column: margin
penguins |>
  count(species, name = "n") |>
  knitr::kable()
```

## Bill shape separates the species

@fig-bill shows that within a species the bill has a characteristic shape, so the species form
clusters.

```{r}
#| label: fig-bill
#| fig-cap: "Bill length versus depth, colored by species."
#| fig-alt: >-
#|   Scatter plot of bill depth against bill length for penguins of three species, each drawn in a
#|   distinct color and shape; the three species form largely separate clusters.
ggplot(penguins, aes(bill_len, bill_dep, color = species, shape = species)) +
  geom_point(alpha = 0.8) +
  scale_color_okabe_ito() +   # color-blind-safe palette; shape also encodes species
  labs(x = "Bill length (mm)", y = "Bill depth (mm)", color = "Species", shape = "Species") +
  theme_minimal(base_size = 12)
```

A compact descriptor of bill shape is the length-to-depth ratio (@eq-ratio):

$$
\text{ratio} = \frac{\text{bill\_len}}{\text{bill\_dep}}
$$ {#eq-ratio}

## Mean measurements

@tbl-summary reports the mean bill measurements and body mass per species.

```{r}
#| label: tbl-summary
#| tbl-cap: "Mean measurements per species."
penguins |>
  summarise(
    n = n(),
    bill_len = mean(bill_len),
    bill_dep = mean(bill_dep),
    body_mass = mean(body_mass, na.rm = TRUE),
    .by = species
  ) |>
  gt() |>
  cols_label(species = "Species", n = "N", bill_len = "Bill length (mm)",
             bill_dep = "Bill depth (mm)", body_mass = "Body mass (g)") |>
  fmt_number(columns = c(bill_len, bill_dep), decimals = 1) |>
  fmt_number(columns = body_mass, decimals = 0)
```

<!-- Part 2 picks up here: add `bibliography:` + `csl:` and a `[@gorman2014]` citation, then a
     References section, then switch to `format: typst` for the branded PDF. -->

## Session {.appendix .unnumbered}

<details>
<summary>Session info</summary>

```{r}
#| label: session
sessionInfo()
```

</details>
